---
title: "Module 2 - Data Wrangling and Validation"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we turn raw clinical and omics tables into an analysis-ready patient-level dataset without corrupting identifiers, outcomes, or molecular measurements?

# Learning Objectives

By the end of this module, learners should be able to:

- inspect table dimensions, column names, and data types,
- validate patient identifiers before joining tables,
- audit missing values and categorical levels,
- join clinical and omics data safely,
- reshape expression data between wide and long formats,
- create analysis-ready derived variables with clear meaning.

# Prerequisites

- Module 0: course orientation and scientific workflow.
- Module 1: R objects, tibbles, functions, and stable file paths.
- Packages: `tidyverse`, `here`.
- Datasets: `data/metabric_clinical_and_expression_data.csv`, `data/clinical.csv`, `data/omics.csv`.

# Setup

```{r packages}
library(tidyverse)
library(here)
```

# Data Input

```{r data-input}
metabric <- read_csv(
  here("data", "metabric_clinical_and_expression_data.csv"),
  show_col_types = FALSE
)

clinical <- read_csv(
  here("data", "clinical.csv"),
  show_col_types = FALSE
)

omics <- read_csv(
  here("data", "omics.csv"),
  show_col_types = FALSE
)
```

# First Table Audit

Before any join or transformation, record table size and missingness.

```{r table-audit}
table_audit <- tibble(
  table = c("metabric", "clinical", "omics"),
  rows = c(nrow(metabric), nrow(clinical), nrow(omics)),
  columns = c(ncol(metabric), ncol(clinical), ncol(omics)),
  missing_values = c(sum(is.na(metabric)), sum(is.na(clinical)), sum(is.na(omics)))
)

table_audit
```

# Column Types

Column types tell you what operations are safe. For example, survival time should be numeric, while treatment and status often become factors for modeling and plotting.

```{r column-types}
clinical_types <- tibble(
  column = names(clinical),
  type = map_chr(clinical, ~ class(.x)[1])
)

clinical_types
```

```{r metabric-types-summary}
metabric_type_summary <- tibble(
  column = names(metabric),
  type = map_chr(metabric, ~ class(.x)[1]),
  missing = map_int(metabric, ~ sum(is.na(.x)))
) %>%
  count(type, sort = TRUE)

metabric_type_summary
```

# Identifier Validation

Identifiers are the join contract. If patient IDs are duplicated, missing, or mismatched, downstream analyses can become invalid.

```{r id-validation}
validate_ids <- function(ids) {
  tibble(
    rows = length(ids),
    unique_ids = n_distinct(ids),
    missing_ids = sum(is.na(ids)),
    duplicated_ids = sum(duplicated(ids))
  )
}

bind_rows(
  clinical = validate_ids(clinical$patientID),
  omics = validate_ids(omics$patientID),
  .id = "table"
)
```

```{r id-set-checks}
id_checks <- tibble(
  check = c(
    "clinical IDs are unique",
    "omics IDs are unique",
    "clinical and omics contain the same patient IDs",
    "clinical and omics are already in the same order"
  ),
  passed = c(
    !anyDuplicated(clinical$patientID),
    !anyDuplicated(omics$patientID),
    setequal(clinical$patientID, omics$patientID),
    identical(clinical$patientID, omics$patientID)
  )
)

id_checks
```

# Safe Joining

Never rely on row order when joining tables. Align by patient ID explicitly.

```{r safe-join}
omics_aligned <- omics %>%
  slice(match(clinical$patientID, patientID))

stopifnot(identical(clinical$patientID, omics_aligned$patientID))

clinical_omics <- clinical %>%
  left_join(omics_aligned, by = "patientID")

dim(clinical_omics)
```

# Outcome and Treatment Variables

Turn coded variables into interpretable factors while keeping the original columns available.

```{r recode-clinical}
clinical_clean <- clinical %>%
  mutate(
    treatment_group = factor(
      treatment,
      levels = c(0, 1),
      labels = c("Standard of care", "Standard + chemotherapy")
    ),
    event_status = factor(
      status,
      levels = c(0, 1),
      labels = c("Censored", "Event")
    )
  )

clinical_clean %>%
  count(treatment_group, event_status, name = "patients")
```

# Missing Data Audit

METABRIC contains clinically meaningful missingness. Missingness should be inspected before deciding whether to drop, impute, or model it.

```{r missingness-audit}
metabric_missing <- metabric %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "column",
    values_to = "missing"
  ) %>%
  mutate(fraction_missing = missing / nrow(metabric)) %>%
  arrange(desc(fraction_missing))

metabric_missing %>%
  filter(missing > 0) %>%
  slice_head(n = 10)
```

# Categorical Level Audit

Categorical variables often hide spelling issues, rare levels, or implicit reference groups.

```{r categorical-audit}
metabric %>%
  count(ER_status, PR_status, HER2_status, name = "patients") %>%
  arrange(desc(patients)) %>%
  slice_head(n = 10)
```

# Reshaping Omics Data

High-dimensional expression tables are often stored wide: one patient per row and one feature per column. Many tidyverse operations are easier in long format.

```{r omics-long}
selected_probes <- names(omics)[2:7]

omics_long <- omics %>%
  select(patientID, all_of(selected_probes)) %>%
  pivot_longer(
    cols = -patientID,
    names_to = "probe_id",
    values_to = "expression"
  )

omics_long %>%
  slice_head(n = 12)
```

Long format makes summaries by feature straightforward.

```{r omics-summary}
omics_long %>%
  group_by(probe_id) %>%
  summarise(
    mean_expression = mean(expression),
    sd_expression = sd(expression),
    .groups = "drop"
  ) %>%
  arrange(desc(sd_expression))
```

# Build a Starter Analysis Table

For later modules, create a small, patient-level analysis table containing clinical outcomes and a few selected expression probes. This is not a final model table; it is a transparent teaching table.

```{r starter-table}
starter_analysis <- clinical_clean %>%
  left_join(
    omics %>% select(patientID, all_of(selected_probes)),
    by = "patientID"
  )

starter_analysis %>%
  select(patientID, treatment_group, time, event_status, all_of(selected_probes)) %>%
  slice_head(n = 8)
```

# Validation Checklist

```{r validation-checklist}
validation_checklist <- tibble(
  check = c(
    "starter table has one row per patient",
    "no missing patient IDs",
    "all clinical patients retained",
    "selected expression probes are numeric",
    "time is non-negative",
    "status values are valid"
  ),
  passed = c(
    nrow(starter_analysis) == n_distinct(starter_analysis$patientID),
    sum(is.na(starter_analysis$patientID)) == 0,
    nrow(starter_analysis) == nrow(clinical),
    all(map_lgl(starter_analysis[selected_probes], is.numeric)),
    all(starter_analysis$time >= 0),
    all(starter_analysis$status %in% c(0, 1))
  )
)

validation_checklist
```

# Interpretation Checkpoint

Before using `starter_analysis`, answer:

- What is one row?
- Which column is the patient identifier?
- Which variables are outcomes?
- Which variables are predictors?
- Which transformations changed interpretation?
- Which validation checks would make you stop and investigate?

# Common Failure Modes

- Joining clinical and omics tables by row order instead of patient ID.
- Dropping rows with missing values before checking what was lost.
- Converting coded variables to factors without preserving their meaning.
- Treating probe IDs as gene symbols.
- Reshaping expression data without checking that values remain numeric.

# Exercises

1. Identify the five METABRIC columns with the highest fraction of missing values.
2. Count patients by `PAM50` subtype and `Vital_status`.
3. Select ten expression probes from `omics.csv`, reshape them to long format, and compute their standard deviations.
4. Create a validation table for METABRIC patient IDs, survival time, and vital status.
5. Explain why `left_join()` is safer than column-binding clinical and omics tables.

# Key Takeaways

- Wrangling is part of the scientific workflow, not administrative cleanup.
- Patient identifiers are the contract between clinical and molecular data.
- Missingness and categorical levels should be audited before analysis.
- Long and wide data formats serve different purposes.
- A clean analysis table should be accompanied by validation checks.

# Reproducibility

```{r session-info}
sessionInfo()
```
