---
title: "Geographic Data in R"
output: html_document
---

# Geographic Data and Mapping

Geographic visualization is a common task in data science. In R, `ggplot2` combined with the `maps` package provides an easy way to visualize geographic data.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(maps)
```

## Static Maps with `ggplot2`

`ggplot2` provides `geom_polygon()` combined with `coord_quickmap()` for plotting spatial objects.

```{r}
# Load world map data
world <- map_data("world")

# Plot world map
ggplot(data = world, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "antiquewhite", color = "gray") +
  coord_quickmap() +
  theme_minimal()
```

## Plotting Points on a Map

We can overlay data points (like cities) onto a map.

```{r}
# Example cities
cities <- tibble(
  city = c("London", "New York", "Tokyo"),
  lat = c(51.5074, 40.7128, 35.6895),
  lon = c(-0.1278, -74.0060, 139.6917)
)

ggplot(data = world, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "antiquewhite", color = "gray") +
  geom_point(data = cities, aes(x = lon, y = lat, group = NULL), color = "red", size = 3) +
  coord_quickmap() +
  theme_bw()
```

## Summary

-   **Data Standard:** Use the `maps` package to get boundary data.
-   **Static Maps:** Use `geom_polygon()` and `coord_quickmap()` in `ggplot2` for reproducible static figures.
