---
title: "R for Biologists | Module 9 — Pathway Enrichment: Gene Ontology and Gene Set Enrichment Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

> **Acknowledgment:** This module is adapted from the RNA-seq workshop by the National Bioinformatics Infrastructure Sweden (NBIS) (https://github.com/NBISweden/workshop-RNAseq).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

Which biological pathways, functional systems, and Gene Ontology (GO) terms are altered in mouse colons during DSS-induced colitis, and what molecular mechanisms do they reveal about the disease?

# Learning Objectives

By the end of this module, learners should be able to:

- explain the conceptual differences between Over-Representation Analysis (ORA) and Gene Set Enrichment Analysis (GSEA), including their input formats and statistical assumptions,
- map Ensembl identifiers to gene symbols programmatically using a local annotation reference,
- perform ORA using the `enrichR` package to query Gene Ontology (GO) databases,
- calculate enrichment significance, adjust for multiple testing (FDR), and visualize ORA results as bar plots,
- rank genes using fold change and p-value metrics for GSEA,
- run GSEA using the `fgsea` package with local MSigDB gene signature GMT files,
- interpret GSEA output (NES, p-values, running enrichment scores) and identify up- and down-regulated pathways,
- write an interpretation paragraph connecting pathway enrichment results to colitis pathophysiology.

# Prerequisites

- Module 2: data wrangling and validation.
- Module 4: statistical reasoning and multiple testing.
- Module 8: RNA-seq count-based differential expression.
- Packages: `tidyverse`, `here`, `enrichR`, `fgsea`, `pheatmap`.
- Datasets: `data/rnaseq_dge_results.csv` (output from Module 8), `data/rnaseq_mouse_genes.txt`, `data/MSigDB_files/`.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(enrichR)
library(fgsea)
library(pheatmap)
```

# Data Input and Mapping

Pathway enrichment databases are usually indexed by human-readable gene symbols (e.g. `TNF`, `IL6`) rather than Ensembl transcript IDs (e.g. `ENSMUSG00000024401`). We must first load our differential expression results and map the Ensembl IDs to gene symbols.

Let's load the DGE results and the gene mapping reference table.

```{r load-data}
dge_path <- here("data", "rnaseq_dge_results.csv")
genes_map_path <- here("data", "rnaseq_mouse_genes.txt")

# Read DGE results (verify file exists, otherwise we'll run a fallback calculation)
if (!file.exists(dge_path)) {
  stop("Differential expression results not found. Please complete Module 8 first.")
}

dge_results <- read_csv(dge_path, show_col_types = FALSE)
genes_map <- read_tsv(genes_map_path, show_col_types = FALSE)
```

Now, map the Ensembl IDs to gene symbols using a left join.

```{r map-genes}
annotated_results <- dge_results %>%
  left_join(
    genes_map %>% select(ensembl_gene_id, external_gene_name),
    by = "ensembl_gene_id"
  ) %>%
  mutate(
    gene_symbol = if_else(is.na(external_gene_name) | external_gene_name == "", ensembl_gene_id, external_gene_name)
  ) %>%
  select(-external_gene_name)

annotated_results %>%
  arrange(padj) %>%
  select(ensembl_gene_id, gene_symbol, log2FoldChange, pvalue, padj) %>%
  dplyr::slice_head(n = 10)
```

# Over-Representation Analysis (ORA)

Over-Representation Analysis evaluates whether the number of significant genes in our list that fall into a specific pathway is larger than would be expected by random chance (modeled using the Hypergeometric distribution).

ORA requires a user-defined threshold to select "significant" genes (e.g. FDR < 0.05 and absolute log2 fold change > 1).

Let's extract the significant up-regulated and down-regulated gene lists as gene symbols.

```{r extract-gene-lists}
sig_up <- annotated_results %>%
  filter(padj < 0.05, log2FoldChange > 1) %>%
  pull(gene_symbol)

sig_down <- annotated_results %>%
  filter(padj < 0.05, log2FoldChange < -1) %>%
  pull(gene_symbol)

# Clean up lists by removing empty values or NAs
sig_up <- sig_up[!is.na(sig_up) & sig_up != ""]
sig_down <- sig_down[!is.na(sig_down) & sig_down != ""]

# Check sizes
tibble(
  sig_up_count = length(sig_up),
  sig_down_count = length(sig_down)
)
```

## Running ORA using enrichR

We query the Gene Ontology Biological Process database (`GO_Biological_Process_2023`).

```{r run-enrichr}
db <- "GO_Biological_Process_2023"

# Query the databases
enrich_up <- enrichr(genes = sig_up, databases = db)
enrich_down <- enrichr(genes = sig_down, databases = db)

# Convert results to tibble
enrich_up_tb <- as_tibble(enrich_up[[db]])
enrich_down_tb <- as_tibble(enrich_down[[db]])
```

Let's view the top 10 enriched pathways for up-regulated genes.

```{r view-enrich-up}
enrich_up_tb %>%
  arrange(P.value) %>%
  select(Term, Overlap, P.value, Adjusted.P.value) %>%
  dplyr::slice_head(n = 10)
```

Let's plot the top 10 enriched pathways for up-regulated genes.

```{r plot-enrich-up}
enrich_up_tb %>%
  arrange(P.value) %>%
  dplyr::slice_head(n = 10) %>%
  mutate(Term = fct_reorder(Term, -P.value)) %>%
  ggplot(aes(x = -log10(P.value), y = Term)) +
  geom_bar(stat = "identity", fill = "#d73027") +
  labs(
    title = "GO Biological Process: Up-regulated in Colitis",
    x = "-Log10 P-value",
    y = NULL
  ) +
  theme_minimal()
```

# Gene Set Enrichment Analysis (GSEA)

A major limitation of ORA is that it depends on an arbitrary significance threshold and discards all information about genes that fall just below the threshold. 

Gene Set Enrichment Analysis (GSEA) does not use a threshold. Instead, it ranks **all** genes in the genome (from most up-regulated to most down-regulated) and calculates a **running enrichment score** to see if the genes in a pathway tend to accumulate at the extremes of the ranked list.

## 1. Preparing the ranked gene vector

We rank our genes by a score that combines the direction of change (sign of log2 fold change) and statistical significance ($-\log_{10}$ p-value). Because pathway databases use uppercase gene symbols, we convert our symbols to uppercase.

```{r prepare-ranked-list}
# Handle duplicate gene symbols by selecting the probe/Ensembl ID with the lowest p-value
ranked_genes <- annotated_results %>%
  filter(!is.na(gene_symbol) & gene_symbol != "") %>%
  mutate(
    rank_score = sign(log2FoldChange) * -log10(pmax(pvalue, .Machine$double.xmin))
  ) %>%
  group_by(gene_symbol) %>%
  dplyr::slice_min(pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(gene_symbol = toupper(gene_symbol))

# Convert to named vector
ranked_vector <- ranked_genes$rank_score
names(ranked_vector) <- ranked_genes$gene_symbol

# Sort decreasing
ranked_vector <- sort(ranked_vector, decreasing = TRUE)
head(ranked_vector)
```

## 2. Load GMT Signature Files

We load the MSigDB Hallmark signatures from our local data folder.

```{r load-signatures}
gmt_path <- here("data", "MSigDB_files", "h.all.v6.2.symbols.gmt.txt")
hallmark_pathways <- gmtPathways(gmt_path)
```

## 3. Running fgsea

We use the `fgsea()` function to perform fast gene set enrichment analysis.

```{r run-fgsea}
fgsea_res <- fgsea(
  pathways = hallmark_pathways,
  stats = ranked_vector,
  minSize = 10,
  maxSize = 500
)

fgsea_res_tb <- as_tibble(fgsea_res) %>%
  arrange(pval)

fgsea_res_tb %>%
  select(pathway, pval, padj, NES, size) %>%
  dplyr::slice_head(n = 10)
```

- **NES (Normalized Enrichment Score)**: Positive scores indicate enrichment at the top of the ranked list (up-regulated in colitis); negative scores indicate enrichment at the bottom (down-regulated).
- **padj**: Benjamini-Hochberg adjusted p-value.

Let's visualize the GSEA results by plotting the NES for all significantly altered Hallmark pathways (FDR < 0.05).

```{r plot-gsea}
fgsea_res_tb %>%
  filter(padj < 0.05) %>%
  mutate(pathway = str_replace(pathway, "^HALLMARK_", "")) %>%
  mutate(pathway = fct_reorder(pathway, NES)) %>%
  ggplot(aes(x = NES, y = pathway, fill = NES > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "#d73027", "FALSE" = "#4575b4"), labels = c("TRUE" = "Up-regulated", "FALSE" = "Down-regulated")) +
  labs(
    title = "GSEA: Hallmark Pathways (FDR < 0.05)",
    x = "Normalized Enrichment Score (NES)",
    y = NULL,
    fill = "Direction"
  ) +
  theme_minimal()
```

## 4. Plotting GSEA Running Enrichment Scores

For a specific pathway of interest, we can plot the running enrichment score. Let's select the top up-regulated pathway.

```{r plot-running-score}
top_pathway <- fgsea_res_tb %>%
  filter(NES > 0) %>%
  arrange(pval) %>%
  dplyr::slice(1) %>%
  pull(pathway)

plotEnrichment(hallmark_pathways[[top_pathway]], ranked_vector) +
  labs(
    title = paste("GSEA Running Enrichment Score:", top_pathway),
    subtitle = "Peak enrichment score represents the degree of coordinate pathway regulation"
  )
```

The vertical black bars along the x-axis represent the positions of the pathway's genes within the ranked genome. The green line shows the cumulative sum running score. The peak score represents the enrichment score (ES).

# Interpretation Checkpoint

Answer these before drawing biological conclusions:

1. Did you map Ensembl IDs to symbols, and how did you handle duplicate symbols?
2. What are the key mathematical differences between ORA and GSEA?
3. In ORA, did you run the enrichment on up- and down-regulated genes separately? (Why is running them together a mistake?)
4. What does the GSEA Normalized Enrichment Score (NES) represent?
5. How does the colitis biological phenotype (tissue damage, immune infiltration) explain both the up-regulated inflammation pathways and down-regulated metabolic pathways?

# Common Failure Modes

- **Mixing up- and down-regulated genes in ORA**: Running ORA on a single combined list of all significant genes hides the direction of change. A pathway can appear significant, but you will not know if it is turned on or off. Always analyze up- and down-regulated lists separately.
- **Ignoring duplicate gene symbols**: Multiple microarray probes or transcript isoforms can map to the same gene symbol. Failing to collapse duplicates when ranking for GSEA creates artificial duplicates that skew the random permutation step.
- **Using an arbitrary rank metric for GSEA without biological justification**: Ranking genes by p-value alone ignores fold change direction, while ranking by fold change alone ignores statistical noise. A combined metric (like `sign(LFC) * -log10(p-value)`) is standard because it balances effect size and statistical confidence.

# Exercises

1. Perform ORA on the list of significant *down-regulated* genes (`sig_down`) using the `GO_Biological_Process_2023` database. Plot the top 10 terms and compare their biological themes with the up-regulated terms.
2. Load the KEGG pathways database signature from the local GMT file (`data/MSigDB_files/c2.cp.kegg.v6.2.symbols.gmt.txt`) using `gmtPathways()`. Run `fgsea()` and identify the top 5 altered KEGG pathways.
3. Compare the enrichment results for the Hallmark pathway **HALLMARK_INFLAMMATORY_RESPONSE** between ORA (from the up-regulated list) and GSEA. Do both methods confirm its activation?
4. Modify the gene ranking metric for GSEA to use the log2 fold change directly (`rank_score = log2FoldChange`) without incorporating the p-value. Run `fgsea()` again and compare the top 5 pathways. Do they differ? Why or why not?

# Key Takeaways

- Pathway enrichment analysis translates gene-level statistical results into biological systems and pathways.
- ORA tests if a discrete list of significant genes is enriched in a pathway using the Hypergeometric test.
- GSEA evaluates coordinate expression shifts across the entire ranked genome, providing higher sensitivity to detect subtle but consistent changes.
- In colitis models, up-regulated genes are associated with immune activation (cytokines, cell migration), while down-regulated genes reflect metabolic and functional collapse of the colon tissue.

# Memorable Rule

Assay decides model.

# Reproducibility

```{r session-info}
sessionInfo()
```
