---
title: "R for Biologists | Module 12 — Reproducible Research: renv, R Markdown, Docker, and Git"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we structure and document a data science analysis so that any biologist on the team — or a future version of us — can reproduce every result exactly, from a fresh clone, six months from now?

# Learning Objectives

By the end of this module, learners should be able to:

- create a well-structured RStudio project with conventional directories,
- use `here` to write portable file paths that survive across machines,
- use `renv` to lock R package versions into a reproducible snapshot,
- write a scripted analysis entrypoint that renders all notebooks,
- parameterize an R Markdown report so different datasets or parameters can be run without editing the code,
- apply `.gitignore` strategy for data, results, and rendered outputs,
- explain the purpose of Docker and containerization for computational reproducibility,
- describe how a CI pipeline can render notebooks automatically on every push.

# Prerequisites

- Module 1: R foundations, reading files, project structure.
- Module 2: data wrangling and validation workflows.
- Packages: `here`, `renv`, `rmarkdown`, `tidyverse`.
- A git installation accessible from the command line.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(renv)
```

# The Reproducibility Problem

Imagine a collaborator opens your analysis folder three months after you wrote it. What do they find?

```{r reproducibility-scenarios}
tribble(
  ~scenario,                       ~outcome,
  "Hardcoded absolute path",       "Breaks on any machine that is not yours",
  "Undocumented package versions", "Different behaviour after a CRAN update",
  "Data edited manually in Excel", "No record of what changed, when, or why",
  "Scripts run in unknown order",  "Results cannot be reproduced without guessing",
  "Results committed to git",      "Version control bloated; reproducibility still depends on re-running scripts"
)
```

Reproducibility is not a style preference. It is a quality control requirement. Every finding in a published analysis must be traceable back to documented code, documented data, and documented computational environment.

# Project Directory Structure

A good project layout separates inputs, code, and outputs.

```{r directory-convention}
cat(
  "project/",
  "  data/          <- read-only source data, never edited by hand",
  "    raw/         <- original files, never modified",
  "    processed/   <- outputs of documented cleaning scripts",
  "  R/             <- reusable functions, not scripts that run analysis",
  "    helpers.R",
  "    checks.R",
  "  course/        <- analysis notebooks (this project)",
  "  scripts/       <- optional entrypoint scripts for larger projects",
  "  outputs/       <- figures and tables produced by scripts",
  "  project_detail/<- planning and implementation documentation",
  "  renv.lock      <- pinned package versions",
  "  DESCRIPTION    <- project metadata and declared dependencies",
  "  .gitignore     <- files git should not track",
  "  README.md      <- how to reproduce the analysis from scratch",
  sep = "\n"
)
```

The key rule: **code lives in version control; raw data does not (unless it is small and static); results are always regenerated from code.**

# Portable Paths with `here`

Absolute paths break when the project moves to a different machine, user account, or operating system.

```{r here-demo}
# here() always starts from the project root, wherever R is called from
here("data", "clinical.csv")
here("outputs", "survival_plot.png")
here("course", "11_survival_analysis_and_omics_integration", "11_survival_analysis_and_omics_integration.Rmd")
```

```{r here-rules}
tribble(
  ~pattern,                  ~problem,                        ~fix,
  "setwd('/home/user/proj')", "Breaks on every other machine", "Use here() instead",
  "read_csv('../data/x.csv')", "Fails when knitting from a subdirectory", "here('data', 'x.csv')",
  "source('helpers.R')",      "Depends on working directory",  "source(here('R', 'helpers.R'))"
)
```

# Dependency Management with `renv`

R packages change. An analysis that runs today may fail in one year when a dependency updates its API, removes a function, or changes default behaviour.

`renv` solves this by recording the exact version of every installed package into a `renv.lock` file.

```{r renv-workflow}
cat(
  "# 1. Initialise renv for the project (run once)",
  "#    renv::init()",
  "#",
  "# 2. After installing or updating packages, snapshot",
  "#    renv::snapshot()",
  "#",
  "# 3. On a fresh clone or new machine, restore the exact environment",
  "#    renv::restore()",
  "#",
  "# 4. Check which packages are out of sync",
  "#    renv::status()",
  sep = "\n"
)
```

```{r renv-lockfile-structure}
cat(
  "renv.lock records, for each package:",
  "  - Package name",
  "  - Version number",
  "  - Repository (CRAN, Bioconductor, GitHub)",
  "  - SHA hash (for GitHub sources)",
  "",
  "This is enough information for renv::restore() to install",
  "exactly the same versions on a different machine.",
  sep = "\n"
)
```

> **Important:** `renv.lock` should be committed to git. The `renv/library/` directory (the actual installed packages) should be in `.gitignore` because it is large and machine-specific.

# The `.gitignore` Strategy

```{r gitignore-rules}
tribble(
  ~file_type,             ~should_be_tracked, ~reason,
  "R scripts and .Rmd",  "Yes",              "These are the primary reproducibility record",
  "renv.lock",           "Yes",              "Pins package versions",
  "DESCRIPTION",         "Yes",              "Declares project dependencies",
  "README.md",           "Yes",              "Entry point for collaborators",
  "data/raw/",           "Only if small and static", "Large data goes in a data repository",
  "outputs/figures/",    "No",               "Always regenerated from scripts",
  "rendered_notebooks/", "No",               "Always regenerated from scripts",
  ".Rhistory, .RData",   "No",               "Machine-specific session debris",
  "renv/library/",       "No",               "Large; restored by renv::restore()"
)
```

# Scripted Analysis Entrypoint

For larger projects, an entrypoint script lets any collaborator reproduce the full analysis with one documented command.

```{r scripted-entrypoint}
cat(
  "# analysis_entrypoint.R — example entrypoint for a project",
  "# Run with: Rscript analysis_entrypoint.R",
  "",
  "library(rmarkdown)",
  "",
  "render_notebook <- function(path, ...) {",
  "  message('Rendering: ', path)",
  "  rmarkdown::render(path, ...)",
  "}",
  "",
  "course_notebooks <- list.files(",
  "  path       = 'course',",
  "  pattern    = '\\\\.Rmd$',",
  "  recursive  = TRUE,",
  "  full.names = TRUE",
  ")",
  "",
  "invisible(lapply(course_notebooks, render_notebook))",
  sep = "\n"
)
```

For production projects, extend this pattern with pass/fail reporting so failures are immediately visible.

# Parameterized R Markdown Reports

Hard-coding dataset names or analysis parameters inside a notebook means copying and editing the notebook each time you want to run it differently. Parameterized reports avoid this.

```{r parameterized-yaml}
cat(
  "---",
  "title: 'Survival Analysis Report'",
  "output: html_document",
  "params:",
  "  data_path: 'data/clinical.csv'",
  "  treatment_var: 'treatment'",
  "  alpha: 0.05",
  "---",
  "",
  "# In the notebook body, use params$data_path, params$treatment_var, etc.",
  "",
  "# Render with different parameters without editing the file:",
  "# rmarkdown::render(",
  "#   'analysis.Rmd',",
  "#   params = list(data_path = 'data/clinical_holdout.csv', alpha = 0.1)",
  "# )",
  sep = "\n"
)
```

Parameterization is especially useful for rerunning analysis on validation cohorts, sensitivity analyses, or different significance thresholds.

# Data Provenance

Every analysis-ready dataset should have a documented lineage.

```{r provenance-template}
tribble(
  ~field,             ~example,
  "Source file",      "data/raw/study_export_2024-01-15.csv",
  "Downloaded from",  "Internal LIMS system, accessed 2024-01-15",
  "Processing script","scripts/01_clean_clinical.R",
  "Output file",      "data/processed/clinical.csv",
  "Key decisions",    "Excluded 3 patients with missing outcome; capped follow-up at 10 years",
  "Analyst",          "A. Ramirez, 2024-01-16",
  "Last verified",    "2024-03-10, re-ran script after R version update, results identical"
)
```

A data dictionary and provenance note prevent the single most common cause of irreproducibility: not knowing which version of the data was analyzed.

# Version Control with Git and GitHub

Git is a version control system that tracks local code history, while GitHub is a web-based hosting service for collaboration, code sharing, and issue tracking. Together, they are the standard infrastructure for modern bioinformatics.

### Core Git Workflow
- **`git clone <url>`**: Copy a remote repository to your local machine.
- **`git branch <name>`**: Create a new development branch to isolate your work (e.g., adding a feature or fixing a bug) without affecting the main code.
- **`git checkout -b <name>`**: Create a new branch and switch to it.
- **`git status`**: Inspect which files have changed, been staged, or are untracked.
- **`git add <file>`**: Stage changes to prepare them for a commit. Use `git add -p` to interactively review changes line-by-line.
- **`git commit -m 'message'`**: Create a permanent snapshot of the staged changes in history.
- **`git push origin <branch>`**: Upload local branch commits to the remote repository.
- **`git pull`**: Fetch and merge changes from the remote repository to your local branch.

### GitHub Collaboration Workflow
1. **Forking**: Copying a public repository to your own GitHub account to make changes independently.
2. **Pull Requests (PR)**: Proposing changes from your fork/branch back to the main repository for review and merging.
3. **Issue Tracking**: Documenting bugs, feature requests, or tasks.

### Commit Message Discipline
Write commit messages that explain *why* a change was made, not just *what* was changed.

```{r git-messages}
tribble(
  ~message,                                             ~evaluation, ~rationale,
  "Update analysis",                                    "BAD",       "Generic and tells the reviewer nothing",
  "Fixed bug in plot",                                  "BAD",       "Doesn't identify the bug or the fix",
  "Fix stratified split: events were leaking into test set", "GOOD",      "Explains the bug and the specific resolution",
  "Add Schoenfeld residual plot for PH assumption check",  "GOOD",      "Clearly declares the new capability and purpose"
)
```

---

# Conda: Managing Non-R System Dependencies

While `renv` is excellent for managing R packages, bioinformatics analyses frequently require command-line programs, compilers, and external libraries (e.g., `samtools`, `bedtools`, `salmon`, `bwa`, or specific Python packages). These cannot be installed using R's `install.packages()`.

**Conda** (and its modern, blazing-fast drop-in C++ replacement **Mamba**) is an open-source package and environment management system that runs on Linux, macOS, and Windows. It allows you to install binaries and their dependencies into isolated environments.

### The Conda Workflow
Conda environments are configured using a YAML file (typically `environment.yml`):

```yaml
name: bioinfo_env
channels:
  - conda-forge
  - bioconda
  - defaults
dependencies:
  - r-base=4.3.1
  - samtools=1.17
  - bedtools=2.31.0
  - salmon=1.10.0
  - python=3.10
  - pip
  - pip:
    - scanpy==1.9.3
```

Using this file, anyone can recreate the exact toolchain with:
```bash
# Create the environment
conda env create -f environment.yml

# Activate the environment
conda activate bioinfo_env
```

---

# Containerization: Docker and Singularity

If you need to freeze the entire computing environment—including the underlying Operating System, system libraries, compilers, Conda environments, and R versions—you use **containers**.

### Docker
Docker builds isolated container images using a recipe called a `Dockerfile`. A typical Dockerfile for an R-based analysis looks like this:

```dockerfile
# Start from a version-controlled R base image
FROM rocker/r-ver:4.3.1

# Install system dependencies required by R packages
RUN apt-get update && apt-get install -y \
    libcurl4-openssl-dev \
    libssl-dev \
    libxml2-dev

# Restore R packages using renv
COPY renv.lock renv.lock
RUN R -e "install.packages('renv'); renv::restore()"

# Copy project files and run analysis
COPY . /project
WORKDIR /project
CMD ["Rscript", "analysis_entrypoint.R"]
```

You can build and run this container on any machine (regardless of whether it runs Linux, macOS, or Windows):
```bash
# Build the container image
docker build -t my-rnaseq-analysis .

# Run the analysis inside the container
docker run --rm my-rnaseq-analysis
```

### Singularity / Apptainer
While Docker is standard for local development, it is typically blocked on High-Performance Computing (HPC) clusters due to security risks (Docker requires root/admin privileges to run). 

In scientific cluster environments, **Singularity** (now **Apptainer**) is used instead. 
- Singularity runs containers as standard user processes without requiring root access.
- It can import Docker images directly:
  ```bash
  singularity pull docker://rocker/r-ver:4.3.1
  ```
- It integrates seamlessly with host cluster filesystems, making it easy to read large genomic files directly from shared storage.

---

# The Modern Bioinformatics Pipeline Ecosystem

For large-scale, automated multi-step analyses, scientists use specialized tools:

1. **`uv`**: A blazing-fast, modern Python package installer and environment resolver written in Rust. It serves as a direct, much faster alternative to `pip` and `conda` for managing Python dependencies in single-cell and deep learning pipelines.
2. **Snakemake**: A Python-based workflow manager that uses "rules" to define inputs, outputs, and execution commands. It automatically determines which steps can run in parallel and handles restarting failed runs from where they stopped.
3. **Nextflow**: A Groovy-based pipeline engine that is widely used for production-grade pipelines (such as `nf-core`). It natively supports containerization (Docker, Singularity) and can run the same pipeline on a laptop, a local HPC scheduler (Slurm, SGE), or cloud providers (AWS, Google Cloud) without changing the pipeline code.

# Continuous Integration for Rendered Notebooks

A CI pipeline can automatically render all notebooks whenever new code is pushed to the repository. This catches render failures before collaborators see broken notebooks.

```{r ci-concept}
cat(
  "A minimal GitHub Actions workflow (.github/workflows/render.yml):",
  "",
  "  on: [push, pull_request]",
  "",
  "  jobs:",
  "    render:",
  "      runs-on: ubuntu-latest",
  "      steps:",
  "        - uses: actions/checkout@v4",
  "        - uses: r-lib/actions/setup-r@v2",
  "          with:",
  "            r-version: '4.3.1'",
  "        - uses: r-lib/actions/setup-renv@v2",
  "        - name: Render all notebooks",
  "          run: Rscript analysis_entrypoint.R",
  sep = "\n"
)
```

CI render checks are the difference between a course that works and a course that used to work.

# Large Data Management

Omics datasets can be gigabytes. Git is not a data repository.

```{r large-data-strategy}
tribble(
  ~data_size,        ~recommended_approach,
  "< 5 MB",          "Commit directly to the git repository if truly static",
  "5 MB - 1 GB",     "Data repository: Zenodo, Figshare, OSF, or institution data store",
  "> 1 GB",          "Cloud storage with signed URLs; or HPC shared filesystem",
  "All sizes",       "Document download URL, version, and checksum in README or data provenance file"
)
```

For this course, `data/clinical.csv` and `data/omics.csv` are small enough to commit. In a real project with full omics data, only the processed, analysis-ready extract should be tracked.

# Putting It Together: Reproducibility Checklist

This checklist applies to any analysis before sharing or publishing.

```{r reproducibility-checklist}
tribble(
  ~item,                                            ~done,
  "All paths use here() or relative paths",        "[ ]",
  "renv.lock committed and up to date",            "[ ]",
  "Raw data documented with source, date, and version", "[ ]",
  "All notebooks render from a fresh R session",   "[ ]",
  "Analysis entrypoint is documented if the project needs one", "[ ]",
  ".gitignore excludes results, renv/library/, .RData", "[ ]",
  "README explains how to reproduce from a clone", "[ ]",
  "sessionInfo() appears in every notebook",       "[ ]",
  "Random seeds are set and documented",           "[ ]",
  "No manually edited intermediate files",         "[ ]"
) %>%
  rename(Checklist = item, Status = done)
```

# Exercises

1. Run `renv::status()` from the project root. Which packages, if any, are not snapshotted?
2. Create a minimal parameterized `.Rmd` that reads a CSV path from `params$data_path` and prints the number of rows. Render it twice with different `params$data_path` values using `rmarkdown::render()` from the console.
3. Add a new entry to `.gitignore` that excludes any file ending in `_cache/`. Confirm with `git status` that cached render files no longer appear as untracked.
4. Write a `README.md` section titled "Reproducing the analysis" that explains the steps from fresh clone to rendered notebooks. Include `renv::restore()` and the command needed for your chosen entrypoint, if one exists.
5. Inspect the current `DESCRIPTION` file. What would you add to make it accurately describe the course? Add at least one missing field and explain your choice.

# Common Failure Modes

- Using `setwd()` or absolute paths that work only on the author's machine.
- Installing packages interactively without snapshotting `renv.lock`, so a collaborator gets different versions.
- Committing rendered HTML files instead of only committing the `.Rmd` source.
- Running analysis scripts out of order and forgetting to document the correct order in the README.
- Editing a CSV in Excel between scripts and not recording the change anywhere.
- Writing results into a version-controlled directory so git history fills with binary diffs.

# Key Takeaways

- Reproducibility is a first-class scientific requirement, not a cleanup step.
- `here` eliminates path fragility; `renv` eliminates package fragility; Docker eliminates OS fragility.
- Every analysis needs a documented entrypoint so any collaborator can re-run it in the correct order.
- A CI render check catches failures before collaborators see them.
- The goal is that a new team member can clone the repository, run one command, and get identical results.

# Memorable Rule

In-sample is not validated.

# Reproducibility

```{r session-info}
sessionInfo()
```
