---
title: "R for Biologists | Module 0 — Scientific Workflow and Course Orientation"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we turn a biological question into a reproducible data science workflow that can be checked, rerun, and interpreted honestly?

# Learning Objectives

By the end of this module, learners should be able to:

- distinguish descriptive, inferential, and predictive questions,
- identify the unit of analysis in a biological dataset,
- explain why data validation comes before modeling,
- run basic checks on the course datasets,
- describe how the capstone connects clinical data, omics data, statistics, machine learning, and reproducible reporting.

# Course Philosophy

This course is not a package tour. Packages matter, but the central skill is choosing a defensible analysis for a biological question.

Every analysis should start with five questions:

1. What is the biological question?
2. What is the unit of analysis?
3. What data do we have, and what assumptions does it encode?
4. What method is appropriate for the question and data?
5. What conclusion is justified, and what uncertainty remains?

# Three Kinds of Data Science Questions

Biological data science work usually falls into one of three categories.

| Question type | Goal | Example |
|---|---|---|
| Descriptive | Summarize what is in the data | What PAM50 subtypes are represented in METABRIC? |
| Inferential | Estimate or test a relationship | Is survival associated with tumor stage after accounting for age? |
| Predictive | Predict outcomes for unseen samples | Can clinical and expression features predict survival risk? |

These question types overlap, but they should not be confused. A model that predicts well does not automatically explain biological mechanism.

# Setup

```{r packages}
library(tidyverse)
library(here)
```

# Course Datasets

The redesigned course uses three main data assets:

- METABRIC clinical and selected expression features.
- A clinical survival table for 614 breast cancer patients.
- A matched microarray expression matrix for the same 614 patients.

```{r data-paths}
metabric_path <- here("data", "metabric_clinical_and_expression_data.csv")
clinical_path <- here("data", "clinical.csv")
omics_path <- here("data", "omics.csv")

file.exists(c(metabric_path, clinical_path, omics_path))
```

# Data Validation First

Before plotting or modeling, verify dimensions, identifiers, missingness, and obvious alignment problems.

```{r load-data}
metabric <- read_csv(metabric_path, show_col_types = FALSE)
clinical <- read_csv(clinical_path, show_col_types = FALSE)
omics <- read_csv(omics_path, show_col_types = FALSE)

dataset_summary <- tibble(
  dataset = c("metabric", "clinical", "omics"),
  rows = c(nrow(metabric), nrow(clinical), nrow(omics)),
  columns = c(ncol(metabric), ncol(clinical), ncol(omics)),
  missing_values = c(sum(is.na(metabric)), sum(is.na(clinical)), sum(is.na(omics)))
)

dataset_summary
```

# Unit of Analysis

The unit of analysis is the thing represented by one row. Getting this wrong leads to invalid statistics and inflated model performance.

```{r unit-of-analysis}
id_summary <- tibble(
  dataset = c("metabric", "clinical", "omics"),
  id_column = c("Patient_ID", "patientID", "patientID"),
  rows = c(nrow(metabric), nrow(clinical), nrow(omics)),
  unique_ids = c(
    n_distinct(metabric$Patient_ID),
    n_distinct(clinical$patientID),
    n_distinct(omics$patientID)
  )
)

id_summary
```

For the main capstone, the unit of analysis is the patient.

# Clinical and Omics Alignment

The clinical and omics files are intended to be paired by `patientID`. We validate that before using them together.

```{r clinical-omics-alignment}
clinical_ids <- clinical$patientID
omics_ids <- omics$patientID

alignment_summary <- tibble(
  check = c(
    "clinical IDs are unique",
    "omics IDs are unique",
    "same number of IDs",
    "same ID set",
    "same ID order"
  ),
  passed = c(
    !anyDuplicated(clinical_ids),
    !anyDuplicated(omics_ids),
    length(clinical_ids) == length(omics_ids),
    base::setequal(clinical_ids, omics_ids),
    identical(clinical_ids, omics_ids)
  )
)

alignment_summary
```

# First Biological Glimpse

The clinical capstone outcome is time-to-event data. `status` indicates whether the event was observed and `time` is the event or censoring time.

```{r clinical-glimpse}
clinical %>%
  count(treatment, status, name = "patients") %>%
  mutate(
    treatment = if_else(treatment == 0, "standard_of_care", "standard_plus_chemotherapy"),
    status = if_else(status == 0, "censored", "event")
  )
```

```{r time-distribution}
clinical %>%
  mutate(
    treatment = factor(
      treatment,
      levels = c(0, 1),
      labels = c("Standard of care", "Standard + chemotherapy")
    )
  ) %>%
  ggplot(aes(x = time, fill = treatment)) +
  geom_histogram(bins = 40, alpha = 0.75, position = "identity") +
  facet_wrap(~treatment, ncol = 1) +
  labs(
    title = "Observed follow-up time by treatment group",
    x = "Time",
    y = "Patients"
  ) +
  theme_minimal()
```

# Interpretation Checkpoint

Before modeling, answer:

- Are rows patients, samples, genes, or measurements?
- Is the outcome continuous, categorical, count-based, or time-to-event?
- Are there missing values or unmatched identifiers?
- Is the goal explanation, prediction, or both?

# Common Failure Modes

- Treating prediction as proof of biological mechanism.
- Modeling before validating identifiers.
- Normalizing or selecting features before splitting data.
- Ignoring censoring in survival data.
- Reporting p-values without effect sizes, confidence intervals, or multiple-testing correction.

# Exercises

1. Count how many METABRIC patients have each `PAM50` subtype.
2. Calculate the number of expression probes in `omics.csv`.
3. Confirm that `clinical.csv` and `omics.csv` still align after sorting both by `patientID`.
4. Write one descriptive, one inferential, and one predictive question for the capstone dataset.

# Key Takeaways

- A good biological data science workflow starts with the question and unit of analysis.
- Data validation is not optional setup work; it is part of the scientific analysis.
- The capstone dataset is patient-level survival data with matched high-dimensional expression features.
- Reproducibility means another person can rerun the analysis and inspect every decision.

# Memorable Rule

IDs before joins.

# Reproducibility

```{r session-info}
sessionInfo()
```
