---
title: "R for Biologists | Module 5 — Machine Learning Foundations with tidymodels"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we build and evaluate predictive models for biological data without leaking information from the test set into training?

# Learning Objectives

By the end of this module, learners should be able to:

- distinguish predictive modeling from statistical inference,
- create leakage-safe train/test splits,
- use `recipes` for preprocessing inside modeling workflows,
- fit regression and classification models with `tidymodels`,
- evaluate regression and classification performance using appropriate metrics,
- use cross-validation and simple tuning,
- explain why model performance does not imply biological causality.

# Prerequisites

- Module 2: data wrangling and validation.
- Module 4: statistical reasoning and experimental design.
- Packages: `tidyverse`, `here`, `tidymodels`, `ranger`, `vip`.
- Dataset: `data/metabric_clinical_and_expression_data.csv`.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(tidymodels)
library(ranger)
library(vip)

tidymodels_prefer()
```

# Data Input and Validation

```{r data-input}
metabric <- read_csv(
  here("data", "metabric_clinical_and_expression_data.csv"),
  show_col_types = FALSE
)

required_columns <- c(
  "Patient_ID",
  "Age_at_diagnosis",
  "Survival_time",
  "Tumour_size",
  "Mutation_count",
  "Cancer_type",
  "ER_status",
  "PAM50",
  "ESR1",
  "PGR",
  "ERBB2",
  "TP53",
  "PIK3CA",
  "GATA3",
  "FOXA1",
  "MLPH"
)

stopifnot(all(required_columns %in% names(metabric)))
stopifnot(!anyDuplicated(metabric$Patient_ID))
```

# Predictive Modeling Is Not Inference

Predictive modeling asks: "How well can this model predict unseen data?"

It does not automatically answer:

- Is this feature causal?
- Is this feature a therapeutic target?
- Will performance hold in another hospital, assay platform, or population?

Those questions need design, validation, and biological follow-up.

# Example 1: Regression

We start with a regression task: predict tumor size from age, mutation count, cancer type, subtype, and selected expression features.

This is a teaching example, not a recommended clinical model.

```{r regression-data}
regression_data <- metabric %>%
  select(
    Patient_ID,
    Tumour_size,
    Age_at_diagnosis,
    Mutation_count,
    Cancer_type,
    PAM50,
    ESR1,
    PGR,
    ERBB2,
    TP53,
    PIK3CA,
    GATA3,
    FOXA1,
    MLPH
  ) %>%
  drop_na(Tumour_size, Age_at_diagnosis, Cancer_type, PAM50)

regression_data %>%
  summarise(
    patients = n(),
    missing_values = sum(is.na(across(everything())))
  )
```

## Train/Test Split

The test set is held back until final evaluation. It must not influence preprocessing, model selection, or tuning.

```{r regression-split}
set.seed(42)
tumor_split <- initial_split(regression_data, prop = 0.8)
tumor_train <- training(tumor_split)
tumor_test <- testing(tumor_split)

dim(tumor_train)
dim(tumor_test)
```

## Baseline Model

A simple baseline helps interpret whether a more complex model is actually useful.

```{r regression-baseline}
baseline_mean <- mean(tumor_train$Tumour_size)

baseline_results <- tumor_test %>%
  transmute(
    Tumour_size,
    .pred = baseline_mean
  )

baseline_metrics <- metrics(baseline_results, truth = Tumour_size, estimate = .pred)
baseline_metrics
```

## Recipe and Workflow

Preprocessing steps go inside the recipe. When resampling, `tidymodels` estimates those steps from each training fold only.

```{r regression-recipe}
tumor_recipe <- recipe(
  Tumour_size ~ Age_at_diagnosis + Mutation_count + Cancer_type + PAM50 +
    ESR1 + PGR + ERBB2 + TP53 + PIK3CA + GATA3 + FOXA1 + MLPH,
  data = tumor_train
) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

linear_spec <- linear_reg() %>%
  set_engine("lm")

linear_workflow <- workflow() %>%
  add_recipe(tumor_recipe) %>%
  add_model(linear_spec)
```

## Fit and Evaluate

```{r regression-fit}
linear_fit <- fit(linear_workflow, data = tumor_train)

linear_results <- predict(linear_fit, tumor_test) %>%
  bind_cols(tumor_test %>% select(Tumour_size))

linear_metrics <- metrics(linear_results, truth = Tumour_size, estimate = .pred)
linear_metrics
```

```{r regression-observed-predicted}
linear_results %>%
  ggplot(aes(x = Tumour_size, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#D1495B", linetype = "dashed") +
  coord_equal() +
  labs(
    title = "Observed vs predicted tumor size",
    x = "Observed tumor size",
    y = "Predicted tumor size"
  ) +
  theme_minimal()
```

# Cross-Validation

Cross-validation estimates how model performance varies across multiple training/validation splits.

```{r regression-cv}
set.seed(42)
tumor_folds <- vfold_cv(tumor_train, v = 5)

cv_results <- fit_resamples(
  linear_workflow,
  resamples = tumor_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

collect_metrics(cv_results)
```

# Example 2: Classification

Now we predict ER status using selected expression and clinical features. This is intentionally easier than many real biomedical prediction problems because ESR1 expression is biologically tied to ER status.

That makes it useful pedagogically, but it also raises a warning: strong predictive features may be proxies for how a label was measured.

```{r classification-data}
classification_data <- metabric %>%
  select(
    Patient_ID,
    ER_status,
    Age_at_diagnosis,
    Tumour_size,
    Mutation_count,
    Cancer_type,
    PAM50,
    ESR1,
    PGR,
    ERBB2,
    TP53,
    PIK3CA,
    GATA3,
    FOXA1,
    MLPH
  ) %>%
  filter(ER_status %in% c("Negative", "Positive")) %>%
  mutate(ER_status = factor(ER_status, levels = c("Negative", "Positive"))) %>%
  drop_na(ER_status)

classification_data %>%
  count(ER_status, name = "patients") %>%
  mutate(fraction = patients / sum(patients))
```

## Stratified Split

For classification, stratify by the outcome so class balance is similar in training and test sets.

```{r classification-split}
set.seed(42)
er_split <- initial_split(classification_data, prop = 0.8, strata = ER_status)
er_train <- training(er_split)
er_test <- testing(er_split)

bind_rows(
  train = er_train %>% count(ER_status),
  test = er_test %>% count(ER_status),
  .id = "split"
)
```

## Classification Recipe

```{r classification-recipe}
er_recipe <- recipe(
  ER_status ~ Age_at_diagnosis + Tumour_size + Mutation_count + Cancer_type + PAM50 +
    ESR1 + PGR + ERBB2 + TP53 + PIK3CA + GATA3 + FOXA1 + MLPH,
  data = er_train
) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

## Logistic Regression

```{r logistic-model}
logistic_spec <- logistic_reg() %>%
  set_engine("glm")

logistic_workflow <- workflow() %>%
  add_recipe(er_recipe) %>%
  add_model(logistic_spec)

logistic_fit <- fit(logistic_workflow, data = er_train)
```

```{r logistic-predictions}
logistic_predictions <- predict(logistic_fit, er_test, type = "prob") %>%
  bind_cols(predict(logistic_fit, er_test, type = "class")) %>%
  bind_cols(er_test %>% select(ER_status))

logistic_predictions %>%
  slice_head(n = 8)
```

## Classification Metrics

Accuracy alone is rarely enough in biomedical classification. Sensitivity, specificity, ROC AUC, and threshold choice often matter.

```{r classification-metrics}
accuracy(logistic_predictions, truth = ER_status, estimate = .pred_class)
sens(logistic_predictions, truth = ER_status, estimate = .pred_class, event_level = "second")
spec(logistic_predictions, truth = ER_status, estimate = .pred_class, event_level = "second")
roc_auc(logistic_predictions, truth = ER_status, .pred_Positive, event_level = "second")
```

```{r confusion-matrix}
conf_mat(logistic_predictions, truth = ER_status, estimate = .pred_class)
```

```{r roc-curve}
logistic_predictions %>%
  roc_curve(truth = ER_status, .pred_Positive, event_level = "second") %>%
  autoplot() +
  labs(title = "ROC curve for ER status classifier") +
  theme_minimal()
```

# Threshold Choice

The default 0.5 probability threshold may not match the scientific or clinical cost of false positives and false negatives.

```{r threshold-example}
threshold_results <- logistic_predictions %>%
  transmute(
    ER_status,
    predicted_positive_30 = factor(
      if_else(.pred_Positive >= 0.30, "Positive", "Negative"),
      levels = c("Negative", "Positive")
    ),
    predicted_positive_70 = factor(
      if_else(.pred_Positive >= 0.70, "Positive", "Negative"),
      levels = c("Negative", "Positive")
    )
  )

tibble(
  threshold = c(0.30, 0.70),
  sensitivity = c(
    sens(threshold_results, ER_status, predicted_positive_30, event_level = "second")$.estimate,
    sens(threshold_results, ER_status, predicted_positive_70, event_level = "second")$.estimate
  ),
  specificity = c(
    spec(threshold_results, ER_status, predicted_positive_30, event_level = "second")$.estimate,
    spec(threshold_results, ER_status, predicted_positive_70, event_level = "second")$.estimate
  )
)
```

# Simple Model Tuning

Random forests can capture nonlinear patterns and interactions. Tuning should happen inside resampling on the training set only.

```{r random-forest-tuning}
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 200
) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

rf_workflow <- workflow() %>%
  add_recipe(er_recipe) %>%
  add_model(rf_spec)

set.seed(42)
er_folds <- vfold_cv(er_train, v = 5, strata = ER_status)

rf_grid <- grid_regular(
  mtry(range = c(2, 8)),
  min_n(range = c(2, 10)),
  levels = 3
)

set.seed(42)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = er_folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc, accuracy)
)

collect_metrics(rf_tune) %>%
  arrange(desc(mean))
```

```{r finalize-rf}
best_rf <- select_best(rf_tune, metric = "roc_auc")

final_rf_workflow <- finalize_workflow(rf_workflow, best_rf)

final_rf_fit <- fit(final_rf_workflow, data = er_train)

rf_predictions <- predict(final_rf_fit, er_test, type = "prob") %>%
  bind_cols(predict(final_rf_fit, er_test, type = "class")) %>%
  bind_cols(er_test %>% select(ER_status))

tibble(
  model = c("logistic regression", "random forest"),
  roc_auc = c(
    roc_auc(logistic_predictions, ER_status, .pred_Positive, event_level = "second")$.estimate,
    roc_auc(rf_predictions, ER_status, .pred_Positive, event_level = "second")$.estimate
  ),
  accuracy = c(
    accuracy(logistic_predictions, ER_status, .pred_class)$.estimate,
    accuracy(rf_predictions, ER_status, .pred_class)$.estimate
  )
)
```

# Variable Importance

Variable importance can help inspect a fitted model, but it does not prove mechanism. Features can be predictive because they are correlated with causal biology, batch, label generation, or other hidden structure.

```{r variable-importance}
final_rf_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 12) +
  labs(title = "Random forest variable importance for ER status prediction")
```

# Leakage Checklist

Before trusting a model, ask:

- Did we split before preprocessing?
- Were imputation, normalization, and feature selection estimated only from training data?
- Did any duplicate or repeated patient appear in both train and test sets?
- Was the test set used during tuning?
- Does any predictor encode the outcome directly?
- Is the test set representative of the intended future data?
- Is the model being interpreted as prediction rather than causation?

# Interpretation Checkpoint

For each model, answer:

- What is the prediction target?
- What is the unit of analysis?
- What predictors were available at prediction time?
- What metric best matches the biological or clinical use case?
- What are the likely leakage risks?
- What validation would be needed before deployment?

# Common Failure Modes

- Normalizing all data before splitting.
- Selecting top genes using the full dataset before cross-validation.
- Reporting only the best tuned result without a final held-out test.
- Using accuracy for imbalanced biomedical outcomes.
- Interpreting feature importance as causal mechanism.
- Ignoring how labels were generated.
- Treating internal validation as external validation.

# Exercises

1. Fit a baseline classifier that always predicts the majority ER status class and compare its accuracy to logistic regression.
2. Remove `ESR1` from the ER status classifier and observe how performance changes.
3. Add `step_corr()` to remove highly correlated numeric predictors and rerun logistic regression.
4. Tune a random forest using `min_n` only and compare ROC AUC to the untuned logistic regression.
5. Write a model card paragraph that states the target, data, metrics, limitations, and leakage risks.

# Key Takeaways

- Predictive modeling evaluates performance on unseen data.
- Splitting before preprocessing is the core anti-leakage habit.
- Recipes and workflows make preprocessing reproducible inside resampling.
- Biomedical classification needs metrics beyond accuracy.
- Tuning belongs inside cross-validation on the training set.
- Feature importance is model behavior, not biological proof.

# Reproducibility

```{r session-info}
sessionInfo()
```
