---
title: "Module 6 - Biomedical Machine Learning"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we build high-dimensional biomedical prediction models while controlling leakage, class imbalance, overfitting, and overinterpretation?

# Learning Objectives

By the end of this module, learners should be able to:

- explain why high-dimensional omics data increases overfitting risk,
- create leakage-safe train/test splits for matched clinical and omics data,
- fit and tune an elastic net model for high-dimensional predictors,
- evaluate imbalanced biomedical classifiers with ROC AUC and PR AUC,
- apply imbalance handling inside a recipe,
- perform training-only feature filtering,
- inspect model calibration and feature importance cautiously,
- write a short model card with limitations.

# Prerequisites

- Module 2: data wrangling and validation.
- Module 5: modeling foundations with `tidymodels`.
- Packages: `tidyverse`, `here`, `tidymodels`, `glmnet`, `ranger`, `themis`, `vip`.
- Datasets: `data/clinical.csv`, `data/omics.csv`.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(tidymodels)
library(glmnet)
library(ranger)
library(themis)
library(vip)

tidymodels_prefer()
```

# Data Input and Validation

The capstone clinical dataset contains time-to-event data. In this module we temporarily create a binary label, "Event" versus "Censored", to teach high-dimensional classification mechanics.

This is a simplification. Censored patients are not "no event forever"; they are patients whose event was not observed during follow-up. A proper survival model is required for final time-to-event analysis.

```{r data-input}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE)
omics <- read_csv(here("data", "omics.csv"), show_col_types = FALSE)

stopifnot(!anyDuplicated(clinical$patientID))
stopifnot(!anyDuplicated(omics$patientID))
stopifnot(setequal(clinical$patientID, omics$patientID))

omics_aligned <- omics %>%
  slice(match(clinical$patientID, patientID))

stopifnot(identical(clinical$patientID, omics_aligned$patientID))
```

# Build the Modeling Table

Do not use `time` as a predictor for this binary classifier. Follow-up time is part of the outcome process and would leak outcome information.

```{r modeling-table}
modeling_data <- clinical %>%
  transmute(
    patientID,
    event_status = factor(
      if_else(status == 1, "Event", "Censored"),
      levels = c("Censored", "Event")
    ),
    treatment_group = factor(
      treatment,
      levels = c(0, 1),
      labels = c("Standard of care", "Standard + chemotherapy")
    )
  ) %>%
  left_join(omics_aligned, by = "patientID")

modeling_data %>%
  count(event_status, name = "patients") %>%
  mutate(fraction = patients / sum(patients))
```

# High-Dimensional Risk

This dataset has many more molecular predictors than a basic clinical model.

```{r dimensionality}
tibble(
  patients = nrow(modeling_data),
  predictors = ncol(modeling_data) - 2,
  expression_probes = ncol(omics) - 1,
  event_fraction = mean(modeling_data$event_status == "Event")
)
```

High-dimensional models can look impressive while learning noise. Validation and leakage control matter more than model complexity.

# Stratified Train/Test Split

The test set is held out until final evaluation.

```{r split}
set.seed(42)
event_split <- initial_split(modeling_data, prop = 0.8, strata = event_status)
event_train <- training(event_split)
event_test <- testing(event_split)

bind_rows(
  train = event_train %>% count(event_status),
  test = event_test %>% count(event_status),
  .id = "split"
)
```

# Baseline Classifier

Always compare against a simple baseline. For imbalanced data, a model can have high accuracy by mostly predicting the majority class.

```{r baseline}
majority_class <- event_train %>%
  count(event_status, sort = TRUE) %>%
  slice(1) %>%
  pull(event_status)

baseline_predictions <- event_test %>%
  transmute(
    event_status,
    .pred_class = factor(majority_class, levels = levels(event_status)),
    .pred_Event = if_else(.pred_class == "Event", 1, 0),
    .pred_Censored = 1 - .pred_Event
  )

tibble(
  metric = c("accuracy", "balanced accuracy", "ROC AUC", "PR AUC"),
  estimate = c(
    accuracy(baseline_predictions, event_status, .pred_class)$.estimate,
    bal_accuracy(baseline_predictions, event_status, .pred_class)$.estimate,
    roc_auc(baseline_predictions, event_status, .pred_Event, event_level = "second")$.estimate,
    pr_auc(baseline_predictions, event_status, .pred_Event, event_level = "second")$.estimate
  )
)
```

# Elastic Net for High-Dimensional Data

Elastic net regularization is useful when predictors are numerous and correlated. It combines lasso-like feature selection with ridge-like shrinkage.

```{r elastic-net-recipe}
elastic_recipe <- recipe(event_status ~ ., data = event_train) %>%
  update_role(patientID, new_role = "id") %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

elastic_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

elastic_workflow <- workflow() %>%
  add_recipe(elastic_recipe) %>%
  add_model(elastic_spec)
```

# Cross-Validation and Tuning

Tuning is performed only on the training set. The test set remains untouched.

```{r elastic-net-tuning}
set.seed(42)
event_folds <- vfold_cv(event_train, v = 5, strata = event_status)

elastic_grid <- grid_regular(
  penalty(range = c(-4, -1)),
  mixture(range = c(0, 1)),
  levels = c(5, 3)
)

set.seed(42)
elastic_tune <- tune_grid(
  elastic_workflow,
  resamples = event_folds,
  grid = elastic_grid,
  metrics = metric_set(roc_auc, pr_auc, accuracy, bal_accuracy)
)

collect_metrics(elastic_tune) %>%
  arrange(.metric, desc(mean)) %>%
  group_by(.metric) %>%
  slice_head(n = 3) %>%
  ungroup()
```

```{r final-elastic-net}
best_elastic <- select_best(elastic_tune, metric = "roc_auc")

final_elastic_workflow <- finalize_workflow(elastic_workflow, best_elastic)
final_elastic_fit <- fit(final_elastic_workflow, data = event_train)

elastic_predictions <- predict(final_elastic_fit, event_test, type = "prob") %>%
  bind_cols(predict(final_elastic_fit, event_test, type = "class")) %>%
  bind_cols(event_test %>% select(patientID, event_status))

tibble(
  metric = c("accuracy", "balanced accuracy", "ROC AUC", "PR AUC"),
  estimate = c(
    accuracy(elastic_predictions, event_status, .pred_class)$.estimate,
    bal_accuracy(elastic_predictions, event_status, .pred_class)$.estimate,
    roc_auc(elastic_predictions, event_status, .pred_Event, event_level = "second")$.estimate,
    pr_auc(elastic_predictions, event_status, .pred_Event, event_level = "second")$.estimate
  )
)
```

# ROC and Precision-Recall Curves

ROC AUC can look reasonable even when the minority class is hard to identify. Precision-recall curves are often more informative for rare biomedical outcomes.

```{r roc-pr-curves}
roc_plot <- elastic_predictions %>%
  roc_curve(event_status, .pred_Event, event_level = "second") %>%
  autoplot() +
  labs(title = "Elastic net ROC curve") +
  theme_minimal()

pr_plot <- elastic_predictions %>%
  pr_curve(event_status, .pred_Event, event_level = "second") %>%
  autoplot() +
  labs(title = "Elastic net precision-recall curve") +
  theme_minimal()

roc_plot
pr_plot
```

# Calibration Check

Calibration asks whether predicted probabilities match observed event fractions. This quick binning plot is a diagnostic, not a full calibration study.

```{r calibration}
calibration_table <- elastic_predictions %>%
  mutate(probability_bin = ntile(.pred_Event, 6)) %>%
  group_by(probability_bin) %>%
  summarise(
    patients = n(),
    mean_predicted_event_probability = mean(.pred_Event),
    observed_event_fraction = mean(event_status == "Event"),
    .groups = "drop"
  )

calibration_table
```

```{r calibration-plot}
calibration_table %>%
  ggplot(aes(x = mean_predicted_event_probability, y = observed_event_fraction)) +
  geom_point(size = 3, color = "#33658A") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "Calibration check for elastic net event classifier",
    x = "Mean predicted event probability",
    y = "Observed event fraction"
  ) +
  theme_minimal()
```

# Interpreting Elastic Net Coefficients

Nonzero coefficients identify model-selected predictors. They do not prove biological causality.

```{r nonzero-coefficients}
elastic_coefficients <- final_elastic_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)", estimate != 0) %>%
  arrange(desc(abs(estimate)))

elastic_coefficients %>%
  slice_head(n = 15)
```

# Imbalance Handling Inside the Workflow

If using sampling methods such as SMOTE, apply them inside the recipe so they happen only inside training folds. Never SMOTE the full dataset before splitting.

To keep rendering fast, this example uses a training-only set of high-variance probes before fitting a random forest. The feature screen is computed from the training set only.

```{r training-only-feature-filter}
probe_columns <- setdiff(names(omics), "patientID")

top_train_probes <- event_train %>%
  select(all_of(probe_columns)) %>%
  summarise(across(everything(), sd)) %>%
  pivot_longer(everything(), names_to = "probe", values_to = "sd") %>%
  arrange(desc(sd)) %>%
  slice_head(n = 100) %>%
  pull(probe)

rf_train <- event_train %>%
  select(patientID, event_status, treatment_group, all_of(top_train_probes))

rf_test <- event_test %>%
  select(patientID, event_status, treatment_group, all_of(top_train_probes))
```

```{r smote-rf}
rf_recipe <- recipe(event_status ~ ., data = rf_train) %>%
  update_role(patientID, new_role = "id") %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_smote(event_status, over_ratio = 0.8)

rf_spec <- rand_forest(
  mtry = 20,
  min_n = 10,
  trees = 200
) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_spec)

set.seed(42)
rf_fit <- fit(rf_workflow, data = rf_train)

rf_predictions <- predict(rf_fit, rf_test, type = "prob") %>%
  bind_cols(predict(rf_fit, rf_test, type = "class")) %>%
  bind_cols(rf_test %>% select(patientID, event_status))

tibble(
  model = c("elastic net", "SMOTE random forest"),
  roc_auc = c(
    roc_auc(elastic_predictions, event_status, .pred_Event, event_level = "second")$.estimate,
    roc_auc(rf_predictions, event_status, .pred_Event, event_level = "second")$.estimate
  ),
  pr_auc = c(
    pr_auc(elastic_predictions, event_status, .pred_Event, event_level = "second")$.estimate,
    pr_auc(rf_predictions, event_status, .pred_Event, event_level = "second")$.estimate
  ),
  balanced_accuracy = c(
    bal_accuracy(elastic_predictions, event_status, .pred_class)$.estimate,
    bal_accuracy(rf_predictions, event_status, .pred_class)$.estimate
  )
)
```

# Random Forest Importance

Feature importance is useful for model inspection, but it is not a ranked list of therapeutic targets.

```{r rf-importance}
rf_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 12) +
  labs(title = "Random forest variable importance")
```

# Model Card

A short model card makes limitations explicit.

| Field | Current model |
|---|---|
| Target | Observed event vs censored status |
| Unit of analysis | Patient |
| Predictors | Treatment group and microarray probe intensities |
| Main risk | Censoring is simplified as binary classification |
| Validation | Internal train/test split only |
| Metrics | ROC AUC, PR AUC, balanced accuracy, calibration bins |
| Not supported | Clinical deployment, causal claims, survival risk estimation |

# Interpretation Checkpoint

Before trusting any biomedical ML model, answer:

- Was feature selection done only inside training data?
- Was the test set untouched until final evaluation?
- Is the outcome label biologically and clinically valid?
- Is class imbalance addressed with metrics and, if needed, training-only sampling?
- Are predictors available at the time prediction would be made?
- Has the model been externally validated?
- Are feature-importance claims stated as prediction, not mechanism?

# Common Failure Modes

- Selecting genes on the full dataset before splitting.
- SMOTE or downsampling before train/test split.
- Treating censored patients as true non-events without caveats.
- Reporting accuracy for imbalanced outcomes.
- Confusing internal cross-validation with external validation.
- Treating elastic net selected features as causal biology.
- Forgetting that assay platform and batch can dominate molecular predictors.

# Exercises

1. Refit the elastic net using `pr_auc` instead of `roc_auc` for model selection. Compare the chosen penalty and mixture.
2. Remove `treatment_group` from the model and observe whether performance changes.
3. Change the number of training-selected probes for the random forest from 100 to 50 and compare PR AUC.
4. Write a leakage checklist for a high-dimensional omics classifier.
5. Write a model card paragraph that explains why this binary event classifier is not a proper survival model.

# Key Takeaways

- High-dimensional biomedical ML is mostly a validation problem.
- Elastic net is a practical baseline for many correlated molecular predictors.
- Imbalanced outcomes need PR AUC, balanced accuracy, and threshold thinking, not accuracy alone.
- Sampling methods must occur inside training resamples.
- Feature selection must be training-only.
- Predictive features are not automatically biological mechanisms.
- Censored survival outcomes should eventually be modeled with survival methods.

# Reproducibility

```{r session-info}
sessionInfo()
```
