---
title: "R for Biologists | Module 11 — Single-Cell RNA-seq: QC, Clustering, and Cell Type Annotation"
output:
  html_document:
    toc: true
    toc_depth: 2
---

> **Acknowledgment:** This module is adapted from the single-cell RNA-seq workshop by the National Bioinformatics Infrastructure Sweden (NBIS) (https://github.com/NBISweden/workshop-scRNAseq).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do the cell type compositions and transcriptional profiles of human Peripheral Blood Mononuclear Cells (PBMCs) differ between healthy individuals and COVID-19 patients at single-cell resolution?

# Learning Objectives

By the end of this module, learners should be able to:

- explain the structural differences between bulk RNA-seq and single-cell RNA-seq data,
- construct a `Seurat` object from raw count matrices and access its core slots,
- calculate and interpret single-cell QC metrics (cell count, library size, gene count, mitochondrial percentage),
- perform library size normalization and identify highly variable genes (HVGs),
- explain the purpose of scaling data and running Principal Component Analysis (PCA),
- run and interpret non-linear dimensionality reduction (UMAP) for visualization,
- perform graph-based Louvain clustering and determine optimal resolution,
- identify cluster markers and annotate clusters with cell type names using canonical marker genes,
- run cell-type-specific differential expression between conditions.

# Prerequisites

- Module 2: data wrangling and validation.
- Module 3: visualization for biology (ggplot2).
- Module 8: bulk RNA-seq differential expression (DESeq2).
- Packages: `Seurat`, `tidyverse`, `here`, `patchwork`, `scales`.
- Datasets: `data/seurat_pbmc_subset.rds`.

# Setup

We load `Seurat` and the standard data wrangling packages.

```{r packages}
library(Seurat)
library(tidyverse)
library(here)
library(patchwork)
library(scales)
```

# 1. Loading the Raw Data and Creating the Seurat Object

In single-cell genomics, a dataset is represented by a cell-by-gene expression matrix (where entries are UMI counts). We start with a subset of 1,500 cells from human PBMCs (balanced between 750 COVID-19 patients and 750 healthy controls).

Let's load our raw count matrix and metadata from the subsetted RDS file.

```{r load-raw}
pbmc_data_path <- here("data", "seurat_pbmc_subset.rds")
if (!file.exists(pbmc_data_path)) {
  stop("PBMC dataset not found. Please verify that the data files exist in the data folder.")
}

# Read the pre-processed RDS object to extract its raw counts and metadata
pbmc_in <- readRDS(pbmc_data_path)
counts <- pbmc_in[["RNA"]]$counts
metadata <- pbmc_in@meta.data %>% 
  select(orig.ident, type)

# Create a fresh Seurat object from raw counts
pbmc <- CreateSeuratObject(counts = counts, meta.data = metadata, project = "PBMC_Covid")
pbmc
```

The Seurat object acts as a container that stores the count matrix, normalized data, dimensionality reductions (like PCA and UMAP), and metadata.

Let's inspect the first 5 cells and 5 genes of the count matrix:

```{r inspect-counts}
pbmc[["RNA"]]$counts[1:5, 1:5]
```

Notice the counts matrix is a sparse matrix (`dgCMatrix`) because most cells do not express most genes. Using a sparse matrix saves massive amounts of memory.

---

# 2. Quality Control (QC) Filtering

Single-cell datasets contain low-quality cells, empty droplets, and doublets (multiple cells captured in one droplet). We calculate standard QC metrics:
1. `nCount_RNA`: Total number of UMIs (transcripts) detected per cell.
2. `nFeature_RNA`: Number of unique genes detected per cell.
3. `percent_mito`: Percentage of reads mapping to mitochondrial genes (high values suggest damaged or dying cells whose cytoplasm leaked out).

```{r calculate-qc}
# Mitochondrial genes start with 'MT-' in humans
pbmc <- PercentageFeatureSet(pbmc, pattern = "^MT-", col.name = "percent_mito")
head(pbmc@meta.data)
```

Let's visualize these QC metrics as a violin plot.

```{r plot-qc-violins, fig.width=10, fig.height=6}
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3, pt.size = 0.1)
```

We filter out cells that have:
- Too few genes detected (e.g., < 200, which are likely empty droplets).
- Extremely high mitochondrial percentage (e.g., > 10%, indicating dying cells).

```{r filter-cells}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & percent_mito < 10)
pbmc
```

---

# 3. Normalization and Feature Selection

To compare cells, we must normalize for differences in sequencing depth (library size) across cells. The default method scales the counts per cell to 10,000 and log-transforms the results.

```{r normalize}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE)
```

Next, we identify the **Highly Variable Features** (genes that vary significantly across cells). These genes contain the most signal for clustering and cell type identity.

```{r hvgs}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000, verbose = FALSE)

# Identify the top 10 most variable genes
top10 <- head(VariableFeatures(pbmc), 10)
top10

# Plot variable features
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

---

# 4. Scaling and Dimensionality Reduction (PCA)

Before running PCA, we scale the data so that each gene has a mean of 0 and a variance of 1. This prevents highly expressed genes from dominating the PCA.

```{r scale}
pbmc <- ScaleData(pbmc, verbose = FALSE)
```

Now, we perform linear dimensionality reduction (PCA) on the highly variable genes.

```{r run-pca}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc), verbose = FALSE)

# Visualize PCA loadings for the first two components
DimPlot(pbmc, reduction = "pca", group.by = "type")
```

To decide how many Principal Components (PCs) to retain for downstream clustering and UMAP, we use an **Elbow Plot**, which ranks PCs by the percentage of variance they explain.

```{r elbow-plot}
ElbowPlot(pbmc, ndims = 20)
```

We look for the "elbow" where the curve flattens out (around 10-15 PCs). We will use the first 12 PCs for downstream steps.

---

# 5. Graph-Based Clustering

Single-cell clustering is graph-based: we build a Shared Nearest Neighbor (SNN) graph of cells based on PC distances, and then partition the graph into clusters using the Louvain algorithm.

The `resolution` parameter controls the granularity of clusters: higher values produce more clusters.

```{r clustering}
pbmc <- FindNeighbors(pbmc, dims = 1:12, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.5, verbose = FALSE)

# Show cluster sizes
table(pbmc$seurat_clusters)
```

---

# 6. Non-linear Dimensionality Reduction (UMAP)

For visualization, we project the high-dimensional PCA space into 2D using Uniform Manifold Approximation and Projection (UMAP).

```{r run-umap}
pbmc <- RunUMAP(pbmc, dims = 1:12, verbose = FALSE)

# Plot UMAP colored by cluster
DimPlot(pbmc, reduction = "umap", label = TRUE)
```

We can also overlay the patient type (healthy control vs. COVID-19) onto the UMAP.

```{r plot-umap-type}
DimPlot(pbmc, reduction = "umap", group.by = "type", cols = c("Covid" = "#d73027", "Ctrl" = "#4575b4"))
```

---

# 7. Cluster Marker Identification and Annotation

To identify what biological cell types each cluster represents, we find the differentially expressed genes (markers) for each cluster compared to all other clusters.

```{r find-all-markers}
# Find markers for all clusters (only positive logFC, and present in at least 25% of cells)
markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, verbose = FALSE)

# Show the top 2 markers for each cluster
markers %>%
  group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 2) %>%
  select(cluster, gene, avg_log2FC, p_val_adj)
```

We look up these markers against canonical PBMC cell type markers:
- **T Cells**: `CD3D`, `CD3E` (Clusters 0, 2, 9)
- **CD14+ Monocytes**: `CD14`, `LYZ` (Clusters 1, 4, 7, 8, 12)
- **B Cells**: `MS4A1`, `CD79A` (Clusters 3, 10)
- **NK Cells**: `GNLY`, `NKG7` (Clusters 5, 6)
- **CD16+ Monocytes**: `FCGR3A` (Cluster 11)

Let's visualize the expression of these marker genes across clusters.

```{r feature-plots, fig.width=10, fig.height=8}
FeaturePlot(pbmc, features = c("CD3D", "LYZ", "MS4A1", "GNLY"), ncol = 2)
```

Using these markers, we assign names to the clusters.

```{r annotate-clusters}
new_identities <- c(
  "0" = "T Cells",
  "1" = "CD14+ Monocytes",
  "2" = "T Cells",
  "3" = "B Cells",
  "4" = "CD14+ Monocytes",
  "5" = "NK Cells",
  "6" = "NK Cells",
  "7" = "CD14+ Monocytes",
  "8" = "CD14+ Monocytes",
  "9" = "T Cells",
  "10" = "B Cells",
  "11" = "CD16+ Monocytes",
  "12" = "CD14+ Monocytes"
)

pbmc <- RenameIdents(pbmc, new_identities)

# Add cell type annotation to metadata
pbmc$cell_type <- Idents(pbmc)

# Plot annotated UMAP
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5)
```

---

# 8. Cell-Type-Specific Differential Expression

We can compare COVID-19 vs. healthy controls within a specific cell type to find disease-responsive genes. Let's compare COVID-19 vs. Ctrl within **CD14+ Monocytes**.

```{r cell-type-dge}
# Subset to CD14+ Monocytes
monocytes <- subset(pbmc, idents = "CD14+ Monocytes")

# Set active identity to type (Covid vs Ctrl)
Idents(monocytes) <- "type"

# Find differentially expressed genes
monocyte_dge <- FindMarkers(monocytes, ident.1 = "Covid", ident.2 = "Ctrl", min.pct = 0.25, logfc.threshold = 0.25, verbose = FALSE)

# Show top differentially expressed genes
monocyte_dge %>%
  rownames_to_column("gene") %>%
  arrange(p_val_adj) %>%
  dplyr::slice_head(n = 10)
```

Observe that interferon-induced genes like `ISG15` and `IFITM3` are highly up-regulated in monocytes of COVID-19 patients, reflecting the active antiviral immune response.

---

# Interpretation Checkpoint

Answer these before drawing biological conclusions:

1. Why are single-cell RNA-seq count matrices extremely sparse, and how does Seurat represent them?
2. What are the three standard QC metrics in single-cell QC, and what biological failures do they identify?
3. Why is scaling data necessary before running PCA?
4. What is the difference between PCA (linear) and UMAP (non-linear) dimensionality reduction? Why do we cluster cells in PC space instead of UMAP space?
5. How does cell-type-specific differential expression differ from bulk RNA-seq differential expression?

# Common Failure Modes

- **Clustering directly on UMAP coordinates**: UMAP is a non-linear visualization technique that distorts distances to project high dimensions into 2D. Clustering must always be performed in the low-dimensional PC space (usually the first 10-20 PCs) to preserve linear relationship distances.
- **Ignoring batch effects and integration needs**: When combining single-cell datasets from different patients or labs, cells often cluster by batch/sample instead of cell type. If this occurs, batch integration methods (e.g., Harmony or Seurat v5 integration) are required.
- **Over-filtering biological signal**: Some cells (e.g., active secretory cells or cardiomyocytes) naturally express very high levels of mitochondrial or ribosomal genes. Applying a blind threshold (like < 5% mitochondrial) can delete entire cell types. Always visualize distribution plots before setting thresholds.

# Exercises

1. Calculate the percentage of ribosomal genes (names starting with `RPS` or `RPL`) in our `pbmc` object using `PercentageFeatureSet()`. Plot ribosomal percentage vs. mitochondrial percentage as a scatter plot.
2. Adjust the clustering resolution parameter in `FindClusters()` from `0.5` to `1.2`. Re-run the clustering and check the number of clusters. Does it split the T-cell population into subgroups?
3. Perform marker gene identification for the NK cell cluster (Cluster 6/NK Cells) compared only to the T-cell clusters. What are the top 5 markers distinguishing NK cells from T cells?
4. Identify the top 5 up-regulated genes in **B Cells** of COVID-19 patients compared to healthy controls. Contrast these hits with the CD14+ Monocyte hits.

# Key Takeaways

- Single-cell RNA-seq allows cell-by-cell transcriptional profiling, revealing cell type heterogeneity that is hidden in bulk RNA-seq.
- Standard Seurat workflows include: cell filtering (QC) -> normalization -> variable gene selection -> PCA -> graph-based clustering -> UMAP visualization -> cell type annotation.
- Canonical marker genes (e.g., `CD3D` for T cells, `MS4A1` for B cells) are used to annotate clusters with biological names.
- By comparing conditions within specific cell types, we can identify cell-type-specific disease responses (e.g., interferon responses in monocytes).

# Memorable Rule

Assay decides model.

# Reproducibility

```{r session-info}
sessionInfo()
```
