---
title: "Week 1 - Assignment Solutions"
output: html_document
---

# Week 1 - assignment solution

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(writexl)
```

```{r}
# Read metabric dataset
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")
```

### Exercises from the Text

**A) Write a query to extract data on patients from cohort 1 with either the highest or second highest tumour stage.**

```{r}
# In R, we can find the top 2 unique tumour stages, then filter
top_stages <- metabric %>%
  drop_na(Tumour_stage) %>%
  distinct(Tumour_stage) %>%
  slice_max(Tumour_stage, n = 2) %>%
  pull(Tumour_stage)

metabric %>%
  filter(Tumour_stage %in% top_stages & Cohort == 1)
```

**D) Use colSums to get a vector of the amount of missing data for each variable.**

```{r}
colSums(is.na(metabric))
```

**E) There is a method that does the opposite of `is.na()`...**

```{r}
# In base R, you can just negate is.na() with !
# !is.na(metabric)
# In tidyr, drop_na() removes NAs
```

### Exercise 1

- What are the different values of the `Integrative Cluster` variable? 
- Can you produce a series containing the number of patients in each cluster?

```{r}
# Different values
unique(metabric$Integrative_cluster)
```

```{r}
# Number of patients in each cluster (proportions)
metabric %>%
  count(Integrative_cluster) %>%
  mutate(proportion = n / sum(n))
```

### Exercise 2

- Read the dataset `metabric_clinical_and_expression_data.csv` and store its summary statistics into a new variable called `metabric_summary`.
- Write your summary statistics object into a file called `metabric_summary.csv`.
- Generate a Tab Separated Value (TSV) file instead.
- Produce an Excel spreadsheet containing summary statistics using `writexl`.

```{r}
# Store summary statistics
# In R, summary() returns a table, but for writing to file, we might want a tidy summary
metabric_summary <- summary(metabric)
print(metabric_summary)

# Better summary for export:
metabric_stats <- metabric %>%
  summarise(across(where(is.numeric), list(mean = ~mean(.x, na.rm=TRUE), sd = ~sd(.x, na.rm=TRUE))))
```

```{r, eval=FALSE}
# Write summary statistics
write_csv(metabric_stats, "metabric_summary.csv")
write_tsv(metabric_stats, "metabric_summary.tsv")
write_xlsx(metabric_stats, "metabric_summary.xlsx")
```

### Exercise 3

- Calculate the mean tumour size of patients grouped by vital status and tumour stage.
- Find the cohort of patients and tumour stage where the average expression of genes TP53 and FOXA1 is highest.
- Do patients with greater tumour size live longer? How about patients with greater tumour stage? How about greater Nottingham_prognostic_index?

```{r}
# Grouped mean tumour size
metabric %>%
  group_by(Vital_status, Tumour_stage) %>%
  summarise(mean_size = mean(Tumour_size, na.rm = TRUE), .groups = "drop")
```

```{r}
# Highest average expression of TP53 and FOXA1
# The python code did `.mean()[['TP53', 'FOXA1']]`, so it just printed the averages
metabric %>%
  group_by(Cohort, Tumour_stage) %>%
  summarise(mean_TP53 = mean(TP53, na.rm = TRUE),
            mean_FOXA1 = mean(FOXA1, na.rm = TRUE), 
            .groups = "drop") %>%
  arrange(desc(mean_TP53), desc(mean_FOXA1))
```

```{r}
# Correlations with Survival Time
metabric_dead <- metabric %>% filter(Vital_status == 'Died of Disease')

# Tumour size correlation (Spearman as requested)
cor(metabric_dead$Tumour_size, metabric_dead$Survival_time, method = "spearman", use = "complete.obs")

# Tumour stage grouped averages
metabric_dead %>%
  group_by(Tumour_stage) %>%
  summarise(mean_surv = mean(Survival_time, na.rm = TRUE), n = n())

# Nottingham Prognostic Index correlation
cor(metabric_dead$Nottingham_prognostic_index, metabric_dead$Survival_time, method = "spearman", use = "complete.obs")
```

### Exercise 4

- Which variables (columns) of the metabric dataset have missing data?
- Find the patients ids who have missing tumour size and/or missing mutation count data. Which cohorts do they belong to?
- For the patients identified to have missing tumour size data for each cohort, calculate the average tumour size of the patients with tumour size data available within the same cohort to fill in the missing data.

```{r}
# Variables with missing data
colSums(is.na(metabric))
```

```{r}
# Patients with missing values and their cohorts
metabric %>%
  filter(is.na(Tumour_size) | is.na(Mutation_count)) %>%
  select(Patient_ID, Cohort)

# Proportions per cohort
metabric %>%
  filter(is.na(Tumour_size) | is.na(Mutation_count)) %>%
  count(Cohort) %>%
  mutate(proportion = n / sum(n))
```

```{r}
# Fill missing tumour size with cohort mean
# In Python, this was done with a manual subset and fill, followed by a suggestion for a for loop.
# In R, dplyr's group_by and mutate makes this natively vectorized and grouped.
metabric_filled <- metabric %>%
  group_by(Cohort) %>%
  mutate(Tumour_size = if_else(is.na(Tumour_size), mean(Tumour_size, na.rm = TRUE), Tumour_size)) %>%
  ungroup()

# Validation
sum(is.na(metabric_filled$Tumour_size))
```
