---
title: "Week 1 - Data Handling"
output: html_document
---

# Week 1 - Data Handling

## Useful References
- [R for Data Science (2nd ed.)](https://r4ds.hadley.nz/) — Wickham, Çetinkaya-Rundel & Grolemund
- [tidyverse documentation](https://www.tidyverse.org/)
- [dplyr cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)

## Course Homepage
All materials for this course can be found on the GitHub repository:
[Data Science for Biologists - R Edition](https://github.com/AstraZeneca/data-science-python-course)

## Installing the Tidyverse
R uses packages to extend its base capabilities. For data science, the **tidyverse** is the gold standard collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

If you are running this on your own machine, you will need to install the package first using:
`install.packages("tidyverse")`

## Learning Objectives
In this session we will cover:
- How to read data into R
- How to explore and summarise data
- How to select rows and columns
- How to filter data conditionally
- How to handle missing data
- How to group and summarise data
- How to merge and concatenate datasets

## Loading the Tidyverse
Once installed, a package must be loaded into your current R session using the `library()` function.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Reading in Data
The `readr` package (part of the tidyverse) provides the `read_csv()` function to read comma-separated values files into a `tibble`, which is the tidyverse's version of a data frame.

```{r}
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")
```

## Exploring Your Data

### A Refresher on Objects, Attributes and Functions
In R, `metabric` is now an object (specifically a `tibble`). We can use various functions to explore it.

```{r}
# Number of rows and columns
dim(metabric)
```

```{r}
# Column names
colnames(metabric)
```

```{r}
# View the first few rows
head(metabric)
```

We can also specify how many rows we want to see:

```{r}
head(metabric, n = 10)
```

The `summary()` function provides useful summary statistics for each column:

```{r}
summary(metabric)
```

You can combine these functions. For example, to get a summary of just the first 10 rows:

```{r}
summary(head(metabric, n = 10))
```

To get a concise summary of the data frame's structure and data types, use `glimpse()`:

```{r}
glimpse(metabric)
```

### Columns
You can extract a single column from a tibble using the `$` operator or `pull()`. This returns a vector.

```{r}
# Using $ operator
head(metabric$Survival_time)
```

```{r}
# Length of the vector
length(metabric$Survival_time)
```

We can compute statistics on this vector:

```{r}
mean(metabric$Survival_time, na.rm = TRUE)
```

Note the `na.rm = TRUE` argument. This tells R to ignore missing values when calculating the mean.

You can also compute means across all numeric columns at once using `summarise(across())`:

```{r}
metabric %>% 
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)))
```

#### Column Creation
New columns can be created using the `mutate()` function. For example, creating a column for current age:

```{r}
metabric <- metabric %>%
  mutate(Age_today = Age_at_diagnosis + Survival_time / 365)
```

## Data Subsetting

In base R, square brackets `[]` are used to select rows and columns. In the tidyverse, we prefer `slice()`, `select()`, and `filter()`.

```{r}
# Select the first three rows
slice(metabric, 1:3)
```

```{r}
# Select specific columns
select(metabric, Survival_time, Tumour_size)
```

```{r}
# Select a range of columns
select(metabric, ESR1:MLPH)
```

### Conditional Subsetting

You can subset a data frame using logical conditions via the `filter()` function.

```{r}
# Select patients with a tumour size greater than or equal to 20
filter(metabric, Tumour_size >= 20)
```

More complicated conditions can be specified using `&` (AND), and `|` (OR).

```{r}
filter(metabric, Tumour_size >= 20 & Survival_time >= 100)
```

Some useful logical operators:
- `==` equal
- `!=` not equal
- `>` greater than
- `>=` greater than or equal to
- `<` less than
- `<=` less than or equal to
- `&` and
- `|` or

The `%in%` operator can help avoid cumbersome OR statements:

```{r}
filter(metabric, Cohort %in% c(1, 4, 5))
```

## Missing Data

### Identifying Missing Data
A missing value is represented in R by `NA`. Typically, functions in R return `NA` if there are missing values unless you specify `na.rm = TRUE`.

```{r}
# Mean will return NA if there are NAs
mean(metabric$Tumour_size)
```

You can identify the location of missing data using `is.na()`.

```{r}
# Count missing values per column
colSums(is.na(metabric))
```

To find which cohort patients belong to when they are missing tumour size data:

```{r}
metabric %>%
  filter(is.na(Tumour_size)) %>%
  pull(Cohort)
```

### Dealing with Missing Data

#### Removal
One option is to remove rows with incomplete information using `drop_na()`.

```{r}
# Drop rows with ANY missing value
drop_na(metabric)
```

```{r}
# Drop rows with missing values in specific columns
drop_na(metabric, Tumour_size, Tumour_stage)
```

#### Replacement
Another option is to fill in missing data. This can be achieved via `replace_na()` or `mutate()`.

```{r}
# Replace missing Tumour_size with the overall mean
mean_size <- mean(metabric$Tumour_size, na.rm = TRUE)

metabric_filled <- metabric %>%
  mutate(Tumour_size = replace_na(Tumour_size, mean_size))

head(metabric_filled)
```

## Grouping
Often you want to compute statistics for different groups within your data. The `group_by()` function combined with `summarise()` makes this incredibly easy.

```{r}
metabric %>%
  group_by(Cohort) %>%
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)))
```

You can group by multiple variables:

```{r}
metabric %>%
  group_by(Cohort, Tumour_stage) %>%
  summarise(mean_survival = mean(Survival_time, na.rm = TRUE), .groups = "drop")
```

## Multiple Data Sets

### Merging Data Sets
Often information is spread across multiple data sets. `dplyr` provides joins (like `left_join`, `inner_join`) to merge them.

```{r}
# Let's create some dummy mutation data to demonstrate merging
set.seed(42)
mutations <- tibble(
  Patient_ID = sample(metabric$Patient_ID, 100),
  Mutation_Status = sample(c("Mutated", "Wildtype"), 100, replace = TRUE)
)
head(mutations)
```

```{r}
# Merge based on Patient_ID
metabric_merged <- inner_join(metabric, mutations, by = "Patient_ID")
head(metabric_merged)
```

### Concatenating Data Sets
Sometimes data sets are split by rows (e.g., different cohorts in different files). We use `bind_rows()` to stack them.

```{r}
# Let's split the data manually to demonstrate
cohort1 <- filter(metabric, Cohort == 1)
cohort2 <- filter(metabric, Cohort == 2)

# Now stack them back together
stacked_cohorts <- bind_rows(cohort1, cohort2)
dim(stacked_cohorts)
```

## Replace

You may want to rename or recode values within a column. `recode()` or `case_match()` (or `if_else()`/`case_when()`) are standard tidyverse ways to achieve this.

```{r}
# Using recode to change Positive/Negative to 1/0
metabric %>%
  mutate(PR_status = recode(PR_status, "Positive" = "1", "Negative" = "0")) %>%
  select(Patient_ID, PR_status) %>%
  head()
```

### Inplace Operations
In Python/pandas, functions often have an `inplace=True` argument. In R, data is immutable by default. To save changes, you *must* explicitly overwrite the variable using `<-`:

```{r}
metabric <- metabric %>%
  mutate(Chemotherapy = recode(Chemotherapy, "YES" = "1", "NO" = "0"))
```

## Summary
In this session we covered:
- Reading data with `read_csv()`
- Exploring data with `head()`, `summary()`, and `glimpse()`
- Subsetting columns with `select()` and rows with `slice()` and `filter()`
- Handling missing data with `is.na()`, `drop_na()`, and `replace_na()`
- Grouping and summarising with `group_by()` and `summarise()`
- Merging data with `inner_join()` and `bind_rows()`
- Recoding values with `mutate()` and `recode()`

## Exercises
Proceed to `week1_solutions.Rmd` for the assignment exercises.
