---
title: "Solutions — Module 2: Data Wrangling and Validation"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
set.seed(42)
```

# Solution 2.1 — Missingness audit

```{r sol-2-1}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE)

colSums(is.na(clinical))

colMeans(is.na(clinical)) > 0.05

clinical <- clinical %>%
  mutate(complete_case = complete.cases(across(everything())))

sum(!clinical$complete_case)
```

All clinical columns show 0 missing values in this dataset, so every patient is a complete case.

# Solution 2.2 — Duplicate detection

```{r sol-2-2}
omics <- read_csv(here("data", "omics.csv"), show_col_types = FALSE)

anyDuplicated(omics$patientID)

stopifnot(!anyDuplicated(omics$patientID))

# Consequence: if rows 10 and 11 were duplicated, a join would create an
# extra row in the merged table and inflate apparent sample size. Summary
# statistics and model training would be biased toward those two patients.
```

# Solution 2.3 — Safe joins

```{r sol-2-3}
inner <- inner_join(clinical, omics, by = "patientID")
nrow(inner)

left  <- left_join(clinical, omics, by = "patientID")
nrow(left)

stopifnot(nrow(left) == nrow(clinical))
stopifnot(!anyDuplicated(left$patientID))
```

`inner_join` and `left_join` both produce 614 rows because every clinical ID matches exactly one omics ID.

# Solution 2.4 — Grouped summaries

```{r sol-2-4}
clinical %>%
  group_by(treatment) %>%
  summarise(
    n          = n(),
    mean_time  = mean(time),
    event_rate = mean(status),
    iqr_time   = IQR(time)
  )

clinical %>%
  count(treatment, status)
```

# Solution 2.5 — Reshaping

```{r sol-2-5}
wide <- tibble(
  patientID = paste0("pt", 1:4),
  probe_A   = c(0.1, 0.4, -0.2, 0.7),
  probe_B   = c(-0.3, 0.2, 0.5, -0.1),
  probe_C   = c(0.8, -0.5, 0.3, 0.2)
)

long <- wide %>%
  pivot_longer(-patientID, names_to = "probe_id", values_to = "expression")
long

long_positive <- long %>% filter(expression > 0)
long_positive

# Pivoting back: rows with expression <= 0 are gone, so the wide form
# will have NA for those probe/patient combinations.
long_positive %>%
  pivot_wider(names_from = probe_id, values_from = expression)
```

The wide table does not match the original because filtering removed negative values. This demonstrates that information lost in long form cannot be recovered without the original data.
