---
title: "Solutions — Module 5: Modeling Foundations with Tidymodels"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(tidymodels)
set.seed(42)
```

```{r load-data}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE) %>%
  mutate(
    status    = factor(status, levels = c(0, 1), labels = c("censored", "event")),
    treatment = factor(treatment)
  )
```

# Solution 5.1 — Stratified splitting

```{r sol-5-1}
split <- initial_split(clinical, prop = 0.75, strata = status)
train <- training(split)
test  <- testing(split)

bind_rows(
  train %>% count(status) %>% mutate(set = "train", prop = n / sum(n)),
  test  %>% count(status) %>% mutate(set = "test",  prop = n / sum(n))
) %>%
  select(set, status, n, prop)
```

Without stratification, small event rates can lead to the test set having no events at all, making AUC meaningless.

# Solution 5.2 — Recipe with preprocessing

```{r sol-5-2}
rec <- recipe(status ~ treatment + time, data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

prepped <- prep(rec, training = train)
juice(prepped) %>% slice_head(n = 5)
```

# Solution 5.3 — Resampling and metric comparison

```{r sol-5-3}
folds <- vfold_cv(train, v = 5, strata = status)

lr_spec <- logistic_reg() %>% set_engine("glm")

lr_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lr_spec)

cv_results <- fit_resamples(
  lr_wf,
  resamples = folds,
  metrics   = metric_set(accuracy, roc_auc, sensitivity)
)

collect_metrics(cv_results)

# Accuracy is misleading because the majority class (censored) dominates.
# A model predicting "censored" for all patients achieves:
mean(train$status == "censored")
```

# Solution 5.4 — Leakage scenario

The leakage step is computing mean expression per probe across **all 614 patients** before splitting. The test set observations contribute to the mean and variance used to select the 50 probes. The model has therefore already "seen" the test set during feature selection.

Leakage-free sequence:

1. Split 80/20.
2. Inside the training fold: compute variance, filter probes, train the model.
3. Apply the training-set filter threshold to the test set probes (do not recompute on the test set).
4. Evaluate on the test set.

In tidymodels: put `step_filter_missing()`, `step_zv()`, and `step_normalize()` inside the recipe — they are estimated only on the training data and applied to the test data.

# Solution 5.5 — Final fit and test evaluation

```{r sol-5-5}
# Using the logistic regression workflow from 5.3
final_fit <- last_fit(lr_wf, split)

collect_metrics(final_fit)

# Test AUC is typically slightly lower than cross-validation AUC because
# CV was run on the training set; any model flexibility adapts to training patterns.
# A large gap suggests overfitting.
```
