---
title: "Solutions — Module 8: RNA-seq Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(DESeq2)
library(enrichR)
library(fgsea)
set.seed(42)
```

```{r load-data}
# Pre-load data for solutions
counts_raw <- read_csv(here("data", "rnaseq_counts_filtered.csv"), show_col_types = FALSE)
metadata   <- read_csv(here("data", "rnaseq_metadata.csv"), show_col_types = FALSE)
genes_map  <- read_tsv(here("data", "rnaseq_mouse_genes.txt"), show_col_types = FALSE)

gene_ids <- counts_raw[[1]]
count_matrix <- counts_raw %>% select(-1) %>% as.matrix()
rownames(count_matrix) <- gene_ids

metadata <- metadata %>%
  mutate(Group = factor(Group, levels = c("day00", "day07")))

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = metadata,
  design = ~ Group
)
```

# Solution 8.1 — Count-based vs. Continuous distributions

Raw RNA-seq count data is made of discrete integers, representing the number of reads mapping to each gene. Standard linear models (like those in `limma` or simple `lm()`) assume that:

1. **Residuals are normally distributed**: Continuous expression data like microarray intensity satisfies this well after log-transformation. However, raw integer count data is highly skewed, discrete, and contains many zeros. Under log-transformation, zero counts become undefined ($-\infty$) or require arbitrary offsets (like $\log(x + 1)$), which distort the statistical relationship.
2. **Homoscedasticity (constant variance)**: Linear regression assumes that the variance of the residuals is constant across all expression levels. In count data, the variance increases with the mean expression level (heteroscedasticity).

The **Negative Binomial** distribution addresses these issues by modeling discrete integer counts directly and accounting for **overdispersion** (where the variance of gene counts is greater than the mean, $\sigma^2 > \mu$). It models the variance as $\sigma^2 = \mu + \alpha\mu^2$, where $\alpha$ is the dispersion parameter.

---

# Solution 8.2 — Manual calculation of size factors

Here is the step-by-step manual calculation of size factors:

```{r sol-8-2}
# 1. Compute the geometric mean for each gene across all 6 samples
# We filter out genes with geometric mean of 0 (at least one zero count)
gm_mean <- function(x) {
  if (any(x == 0)) return(0)
  exp(mean(log(x)))
}
geom_means <- apply(count_matrix, 1, gm_mean)

# Filter count matrix and geom_means to non-zero genes
non_zero_idx <- geom_means > 0
count_matrix_filtered <- count_matrix[non_zero_idx, ]
geom_means_filtered <- geom_means[non_zero_idx]

# 2. Divide each gene's raw count by its geometric mean to create the ratio matrix
ratios <- count_matrix_filtered / geom_means_filtered

# 3. Take the median ratio for each sample as the size factor
manual_size_factors <- apply(ratios, 2, median)
manual_size_factors

# 4. Compare with DESeq2 size factors
dds_temp <- estimateSizeFactors(dds)
deseq_size_factors <- sizeFactors(dds_temp)
deseq_size_factors

# Check if they are identical (within rounding error)
all.equal(manual_size_factors, deseq_size_factors)
```

Let's plot the size factors against the library size (column sums of the raw counts).

```{r sol-8-2-plot}
library_sizes <- colSums(count_matrix)
plot_df <- tibble(
  Sample = names(manual_size_factors),
  SizeFactor = manual_size_factors,
  LibrarySize = library_sizes
)

ggplot(plot_df, aes(x = LibrarySize, y = SizeFactor, label = Sample)) +
  geom_point(size = 3, color = "darkblue") +
  geom_text(vjust = -0.5, hjust = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", color = "grey") +
  labs(
    title = "Size Factors vs. Library Sizes",
    x = "Total Raw Reads (Library Size)",
    y = "Calculated Size Factor"
  ) +
  theme_minimal()
```

**Explanation**: There is a strong positive linear relationship. Size factors primarily correct for differences in sequencing depth (library size). However, because they are calculated using medians, they are robust to a small number of extremely highly expressed genes skewing the total count.

---

# Solution 8.3 — Differential expression parameters

## Part 1: Setting alpha = 0.10

```{r sol-8-3-1}
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
dds <- nbinomWaldTest(dds)

res_alpha_10 <- results(dds, name = "Group_day07_vs_day00", alpha = 0.10)
summary(res_alpha_10)
```

With `alpha = 0.10`, the FDR significance threshold is relaxed, resulting in a higher number of significant genes.

## Part 2: Disabling independent filtering

```{r sol-8-3-2}
res_no_filt <- results(dds, name = "Group_day07_vs_day00", alpha = 0.05, independentFiltering = FALSE)
summary(res_no_filt)
```

**Explanation**: Disabling independent filtering increases the number of genes tested (from ~12,000 to all genes). Low-count genes have very little statistical power to show true differential expression due to high relative noise. By including them in the multiple testing correction, we increase the penalty (larger $M$ in BH correction), which increases the adjusted p-values of moderately expressed genes, thereby reducing overall statistical power.

## Part 3: lfcThreshold vs. Post-hoc filtering

```{r sol-8-3-3}
# Method A: Post-hoc filtering
res_default <- results(dds, name = "Group_day07_vs_day00", alpha = 0.05)
sig_post_hoc <- as.data.frame(res_default) %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)
nrow(sig_post_hoc)

# Method B: Setting lfcThreshold = 1 in results()
res_lfc_1 <- results(dds, name = "Group_day07_vs_day00", alpha = 0.05, lfcThreshold = 1)
summary(res_lfc_1)
```

**Explanation**: Post-hoc filtering tests the null hypothesis $H_0: \text{LFC} = 0$, and then filters genes with large fold changes. Setting `lfcThreshold = 1` changes the null hypothesis to $H_0: |\text{LFC}| \le 1$. The Wald test now evaluates whether the evidence strongly supports a fold change *strictly greater* than 2-fold. This is a much more conservative test, resulting in a smaller list of genes that are robustly and significantly altered.

---

# Solution 8.4 — Visualizing top gene expressions

Let's find the top 4 genes and plot them.

```{r sol-8-4}
# Map genes to get symbols
annotated_results <- as.data.frame(res_default) %>%
  rownames_to_column("ensembl_gene_id") %>%
  left_join(genes_map, by = "ensembl_gene_id") %>%
  mutate(gene_symbol = if_else(is.na(external_gene_name) | external_gene_name == "", ensembl_gene_id, external_gene_name))

top_4_genes <- annotated_results %>%
  arrange(padj) %>%
  slice_head(n = 4)

top_4_ids <- top_4_genes$ensembl_gene_id
top_4_symbols <- top_4_genes$gene_symbol

# Extract normalized counts for these genes
norm_counts <- counts(dds, normalized = TRUE)[top_4_ids, ] %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  pivot_longer(-ensembl_gene_id, names_to = "SampleName", values_to = "Counts") %>%
  left_join(metadata, by = "SampleName") %>%
  left_join(top_4_genes %>% select(ensembl_gene_id, gene_symbol), by = "ensembl_gene_id")

# Plot using ggplot2
ggplot(norm_counts, aes(x = Group, y = Counts, color = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, size = 3) +
  facet_wrap(~ gene_symbol, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Normalized Expression of Top 4 DE Genes",
    x = "Treatment Group",
    y = "Normalized Read Counts (log10 scale)"
  ) +
  theme_minimal()
```

**Explanation**: Scaling the y-axis to a logarithmic scale (`scale_y_log10()`) is helpful because gene expression levels span several orders of magnitude, and variance is proportional to the mean. Log-scaling stabilizes the visual spread of points across samples with high and low expression.

---

# Solution 8.5 — Pathway enrichment and comparison

Let's run ORA on the down-regulated genes.

```{r sol-8-5}
sig_down <- annotated_results %>%
  filter(padj < 0.05, log2FoldChange < -1) %>%
  pull(gene_symbol)
sig_down <- sig_down[!is.na(sig_down) & sig_down != ""]

dbs <- "GO_Biological_Process_2023"
enrich_down <- enrichr(genes = sig_down, databases = dbs)
enrich_down_tb <- as_tibble(enrich_down[[dbs]])

# Top 5 down-regulated pathways
enrich_down_tb %>%
  arrange(P.value) %>%
  select(Term, Overlap, P.value, Adjusted.P.value) %>%
  slice_head(n = 5)
```

**Biological Comparison**:
- **Up-regulated pathways** are dominated by immune responses, leukocyte migration, inflammatory responses, and cytokine production, reflecting the infiltration of immune cells and active inflammation in the colon.
- **Down-regulated pathways** are dominated by metabolic processes (fatty acid oxidation, small molecule metabolism) and ion transport. This reflects the loss of normal colon epithelial cells and functional tissue shutdown (metabolic dysfunction) due to severe tissue damage.

## GSEA Negative NES Interpretation

A negative Normalized Enrichment Score (NES) means that the genes in that pathway are collectively ranked toward the bottom of the ranked list (i.e. they are down-regulated in day 07 colitis compared to day 00 control).

Let's find one Hallmark pathway that is significantly down-regulated.

```{r sol-8-5-gsea}
gmt_path <- here("data", "MSigDB_files", "h.all.v6.2.symbols.gmt.txt")
hallmark_pathways <- gmtPathways(gmt_path)

ranked_genes <- annotated_results %>%
  filter(!is.na(gene_symbol) & gene_symbol != "") %>%
  mutate(rank_score = sign(log2FoldChange) * -log10(pmax(pvalue, .Machine$double.xmin))) %>%
  group_by(gene_symbol) %>%
  slice_min(pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(gene_symbol = toupper(gene_symbol))

ranked_vector <- ranked_genes$rank_score
names(ranked_vector) <- ranked_genes$gene_symbol
ranked_vector <- sort(ranked_vector, decreasing = TRUE)

fgsea_res <- fgsea(pathways = hallmark_pathways, stats = ranked_vector, minSize = 10, maxSize = 500)

as_tibble(fgsea_res) %>%
  filter(padj < 0.05, NES < 0) %>%
  arrange(pval) %>%
  select(pathway, pval, padj, NES, size)
```

A common down-regulated pathway is **HALLMARK_XENOBIOTIC_METABOLISM** or **HALLMARK_FATTY_ACID_METABOLISM**, which confirms that normal colonic metabolic function is severely suppressed during active colitis.
