---
title: "Solutions — Module 8: Survival Analysis and Omics Integration"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(survival)
library(survminer)
library(glmnet)
set.seed(42)
```

```{r load-data}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE)
omics    <- read_csv(here("data", "omics.csv"),    show_col_types = FALSE)

omics_aligned <- omics %>% slice(match(clinical$patientID, patientID))
stopifnot(identical(clinical$patientID, omics_aligned$patientID))

expression_matrix <- omics_aligned %>% select(-patientID) %>% as.matrix()
rownames(expression_matrix) <- omics_aligned$patientID
```

# Solution 8.1 — Censoring scenarios

1. **Event at 730 days** — observed event, status = 1, time = 730.
2. **Alive at study end (day 1095)** — administrative censoring, status = 0, time = 1095.
3. **Moved abroad, last known alive day 400** — loss to follow-up; treated as censoring at day 400, status = 0, time = 400. This is potentially informative censoring if moving is related to health.
4. **Withdrew consent day 200** — also censored at day 200, status = 0, time = 200. Withdrawal may be informative if driven by disease progression.

Cases 2 and 3 are administrative censoring (independent of prognosis). Cases 3 and 4 could be informative censoring.

# Solution 8.2 — Kaplan-Meier interpretation

```{r sol-8-2}
surv_data <- clinical %>%
  mutate(treatment = factor(
    if_else(treatment == 1, "standard_plus_chemo", "standard"),
    levels = c("standard", "standard_plus_chemo")
  ))

km_fit <- survfit(Surv(time, status) ~ treatment, data = surv_data)

# Survival at 2 years
summary(km_fit, times = 2)$surv

# Median survival
km_fit
```

A significant log-rank p-value does not imply causal efficacy when treatment assignment is not randomised: patients in the chemotherapy group may differ systematically in disease severity, age, or centre — any of which could drive survival differences.

# Solution 8.3 — Cox model extension

```{r sol-8-3}
# Including time directly as a predictor is invalid: time is part of the outcome
# Surv(time, status). It cannot also be a covariate. The corrected model:
cox_fit <- coxph(Surv(time, status) ~ treatment, data = surv_data)
summary(cox_fit)

ph_test <- cox.zph(cox_fit)
print(ph_test)
```

If the Schoenfeld p-value is large (> 0.05), there is no evidence against proportional hazards. A small p-value would suggest the treatment effect changes over the follow-up period.

# Solution 8.4 — Validation split for omics Cox

```{r sol-8-4}
set.seed(42)
n <- nrow(clinical)
event_idx    <- which(clinical$status == 1)
censored_idx <- which(clinical$status == 0)

train_idx <- c(
  sample(event_idx,    size = round(0.7 * length(event_idx))),
  sample(censored_idx, size = round(0.7 * length(censored_idx)))
)
test_idx  <- setdiff(seq_len(n), train_idx)

train_clinical <- clinical[train_idx, ]
test_clinical  <- clinical[test_idx,  ]

x_train <- expression_matrix[train_clinical$patientID, ]
x_test  <- expression_matrix[test_clinical$patientID,  ]

y_train <- Surv(train_clinical$time, train_clinical$status)
y_test  <- Surv(test_clinical$time,  test_clinical$status)

cv_cox <- cv.glmnet(x_train, y_train, family = "cox", alpha = 0.5,
                    nfolds = 5, type.measure = "C")

lp_train <- as.numeric(predict(cv_cox, x_train, s = cv_cox$lambda.min, type = "link"))
lp_test  <- as.numeric(predict(cv_cox, x_test,  s = cv_cox$lambda.min, type = "link"))

c_train <- survConcordance(y_train ~ lp_train)$concordance
c_test  <- survConcordance(y_test  ~ lp_test )$concordance

tibble(
  dataset = c("Training (in-sample)", "Test (held-out)"),
  c_index = round(c(c_train, c_test), 3),
  note    = c("Optimistic — model was fit on this data", "Honest estimate of generalisability")
)
```

The gap between training and test C-index is the overfitting gap. Module 8 reported the in-sample C-index; this exercise demonstrates why that figure should never be used as an estimate of clinical utility.

# Solution 8.5 — Risk group reporting (written answer)

A median-split risk group derived from the same data used to fit the model cannot be presented as publishable evidence for a clinical biomarker for the following reasons.

First, the elastic net model was trained on the full cohort. The linear predictor (risk score) was produced by a model that already used the outcome information from every patient, including the ones now labelled "high risk." Any separation visible in the Kaplan-Meier plot is partly a reflection of this circular process.

Second, the cutpoint — the median of the risk scores — was chosen post hoc from the same data. Had a different quantile been tried, a different separation would appear. Choosing the cutpoint that maximises separation inflates the apparent clinical signal beyond what would be observed in a new cohort.

Third, the study reports in-sample performance with no external validation. A clinical biomarker must demonstrate that it identifies risk groups in patients the model has never seen.

Fourth, statistical association in an observational cohort does not establish that the biomarker-defined risk groups should receive different treatments. Clinical utility requires a prospective trial.

Fifth, even if replication is achieved, the probe identifiers in this dataset have not been annotated to gene symbols, and the relationship between probe expression, gene function, and disease biology is unknown.
