---
title: "Solutions - Module 9: Translatome Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(anota2seq)
set.seed(42)
```

# Setup

```{r recreate-analysis-objects}
raw_counts <- read_tsv(here("data", "GSE134606_rawCounts.tsv.gz"), show_col_types = FALSE)

sample_metadata <- tribble(
  ~raw_column,   ~condition, ~replicate, ~assay,
  "P12959_1001", "shCtrl",   1,          "total",
  "P12959_1002", "shCtrl",   1,          "polysome",
  "P12959_1003", "shCtrl",   2,          "total",
  "P12959_1004", "shCtrl",   2,          "polysome",
  "P12959_1005", "shCtrl",   3,          "total",
  "P12959_1006", "shCtrl",   3,          "polysome",
  "P12959_1007", "shCtrl",   4,          "total",
  "P12959_1008", "shCtrl",   4,          "polysome",
  "P12959_1009", "shERa",    1,          "total",
  "P12959_1010", "shERa",    1,          "polysome",
  "P12959_1011", "shERa",    2,          "total",
  "P12959_1012", "shERa",    2,          "polysome",
  "P12959_1013", "shERa",    3,          "total",
  "P12959_1014", "shERa",    3,          "polysome",
  "P12959_1015", "shERa",    4,          "total",
  "P12959_1016", "shERa",    4,          "polysome"
) %>%
  mutate(
    condition = factor(condition, levels = c("shCtrl", "shERa")),
    assay = factor(assay, levels = c("total", "polysome")),
    pair_id = str_c(condition, "_rep", replicate),
    clean_sample = pair_id
  )

count_columns <- setdiff(names(raw_counts), "geneSymbol")
stopifnot(base::setequal(count_columns, sample_metadata$raw_column))

counts_clean <- raw_counts %>%
  filter(!is.na(geneSymbol), geneSymbol != "") %>%
  group_by(geneSymbol) %>%
  summarise(across(all_of(count_columns), ~sum(.x, na.rm = TRUE)), .groups = "drop")

count_matrix <- counts_clean %>%
  column_to_rownames("geneSymbol") %>%
  as.matrix()

storage.mode(count_matrix) <- "integer"

total_metadata <- sample_metadata %>%
  filter(assay == "total") %>%
  arrange(condition, replicate)

polysome_metadata <- sample_metadata %>%
  filter(assay == "polysome") %>%
  arrange(condition, replicate)

dataT <- count_matrix[, total_metadata$raw_column, drop = FALSE]
dataP <- count_matrix[, polysome_metadata$raw_column, drop = FALSE]

colnames(dataT) <- total_metadata$clean_sample
colnames(dataP) <- polysome_metadata$clean_sample

gene_qc <- tibble(
  gene = rownames(dataT),
  keep_any_signal = rowSums(dataT) + rowSums(dataP) > 0
)

dataT_filtered <- dataT[gene_qc$keep_any_signal, , drop = FALSE]
dataP_filtered <- dataP[gene_qc$keep_any_signal, , drop = FALSE]

phenoVec <- total_metadata$condition
names(phenoVec) <- total_metadata$clean_sample

log_cpm <- function(mat) {
  lib_size <- colSums(mat)
  cpm <- sweep(mat, 2, lib_size / 1e6, "/")
  log2(cpm + 1)
}
```

# Solution 9.1 - Verify paired sample order

```{r sol-9-1}
stopifnot(identical(total_metadata$pair_id, polysome_metadata$pair_id))

bad_polysome_metadata <- polysome_metadata %>%
  arrange(desc(raw_column))

identical(total_metadata$pair_id, bad_polysome_metadata$pair_id)
```

The failed check matters because `anota2seq` compares each total RNA column with the corresponding polysome RNA column. If `shCtrl_rep1` total RNA is paired with `shERa_rep4` polysome RNA, the model is mathematically valid but biologically false.

# Solution 9.2 - Library size and zero-count diagnostics

```{r sol-9-2}
paired_library_sizes <- tibble(
  sample = colnames(dataT),
  condition = total_metadata$condition,
  replicate = total_metadata$replicate,
  total_library_size = colSums(dataT),
  polysome_library_size = colSums(dataP)
)

paired_library_sizes

paired_library_sizes %>%
  ggplot(aes(x = total_library_size, y = polysome_library_size, color = condition)) +
  geom_point(size = 3) +
  geom_text(aes(label = replicate), nudge_y = 100000, show.legend = FALSE) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Paired Library Sizes",
    x = "Total RNA counts",
    y = "Polysome RNA counts"
  ) +
  theme_minimal()

zero_summary <- tibble(
  gene = rownames(dataT),
  zero_all_total = rowSums(dataT) == 0,
  zero_all_polysome = rowSums(dataP) == 0
) %>%
  summarise(
    zero_all_total = sum(zero_all_total),
    zero_all_polysome = sum(zero_all_polysome),
    zero_all_both = sum(zero_all_total & zero_all_polysome)
  )

zero_summary
```

Zero-count genes can mean true absence, low expression below detection, annotation problems, or insufficient sequencing depth. They should be filtered or handled explicitly because they cannot support reliable regression.

# Solution 9.3 - Naive translational efficiency score

```{r sol-9-3}
te_score <- log_cpm(dataP_filtered) - log_cpm(dataT_filtered)

condition_lookup <- tibble(
  sample = colnames(te_score),
  condition = total_metadata$condition
)

delta_te <- as_tibble(te_score, rownames = "gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "te_score") %>%
  left_join(condition_lookup, by = "sample") %>%
  group_by(gene, condition) %>%
  summarise(mean_te = mean(te_score), .groups = "drop") %>%
  pivot_wider(names_from = condition, values_from = mean_te) %>%
  mutate(delta_te = shERa - shCtrl) %>%
  arrange(desc(abs(delta_te)))

delta_te %>%
  dplyr::slice_head(n = 10)
```

This ranking is useful for exploration because it highlights genes whose polysome-to-total relationship changes. It is not enough for a claim because ratio-style TE scores can induce spurious correlation and do not run the APV model checks used by `anota2seq`.

# Solution 9.4 - Interpret regulation modes

```{r sol-9-4, eval=FALSE}
ads <- anota2seqDataSetFromMatrix(
  dataP = dataP_filtered,
  dataT = dataT_filtered,
  phenoVec = phenoVec,
  dataType = "RNAseq",
  filterZeroGenes = TRUE,
  normalize = TRUE,
  transformation = "TMM-log2",
  varCutOff = NULL
)

contrast_matrix <- matrix(
  c(-1, 1),
  nrow = length(levels(phenoVec)),
  ncol = 1,
  dimnames = list(levels(phenoVec), "shERa_vs_shCtrl")
)

thresholds <- list(
  minSlopeTranslation = -1,
  maxSlopeTranslation = 2,
  minSlopeBuffering = -2,
  maxSlopeBuffering = 1,
  deltaPT = log2(1.2),
  deltaTP = log2(1.2),
  deltaP = log2(1.2),
  deltaT = log2(1.2),
  maxPAdj = 0.15
)

ads <- anota2seqRun(
  Anota2seqDataSet = ads,
  contrasts = contrast_matrix,
  thresholds = thresholds,
  correctionMethod = "BH",
  performQC = TRUE,
  performROT = TRUE,
  onlyGroup = FALSE
)

reg_modes <- anota2seqGetOutput(
  ads,
  output = "regModes",
  selContrast = 1,
  analysis = "buffering",
  getRVM = TRUE
) %>%
  as_tibble(rownames = "gene")

reg_modes %>%
  count(singleRegMode, sort = TRUE)

reg_modes %>%
  filter(!is.na(singleRegMode)) %>%
  group_by(singleRegMode) %>%
  dplyr::slice_min(apvRvmPAdj, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(gene, singleRegMode, apvEff, apvRvmPAdj)
```

Example cautious interpretation: "This gene is classified as translation-associated under `shERa` knockdown, meaning its polysome-associated abundance changes beyond what is explained by total RNA in this model."

# Solution 9.5 - Tighten the FDR threshold

```{r sol-9-5, eval=FALSE}
thresholds_strict <- thresholds
thresholds_strict$maxPAdj <- 0.05

ads_strict <- anota2seqRun(
  Anota2seqDataSet = ads,
  contrasts = contrast_matrix,
  thresholds = thresholds_strict,
  correctionMethod = "BH",
  performQC = TRUE,
  performROT = TRUE,
  onlyGroup = FALSE
)

reg_modes_strict <- anota2seqGetOutput(
  ads_strict,
  output = "regModes",
  selContrast = 1,
  analysis = "buffering",
  getRVM = TRUE
) %>%
  as_tibble(rownames = "gene")

bind_rows(
  reg_modes %>% count(singleRegMode) %>% mutate(threshold = "FDR 0.15"),
  reg_modes_strict %>% count(singleRegMode) %>% mutate(threshold = "FDR 0.05")
) %>%
  arrange(singleRegMode, threshold)
```

A stricter FDR threshold reduces false positives but may lose true, smaller effects. The right threshold depends on whether the analysis is exploratory or confirmatory.
