---
title: "Solutions — Module 10: Single-Cell RNA-seq Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(Seurat)
library(tidyverse)
library(here)
library(ggplot2)
```

## Setup and Object Recovery

We load the data, create the Seurat object, and run the standard pipeline up to PCA, clustering, and cell type annotation, as described in the tutorial.

```{r prep-seurat}
# Load the downsampled dataset
pbmc_in <- readRDS(here("data", "seurat_pbmc_subset.rds"))
counts <- pbmc_in[["RNA"]]$counts
metadata <- pbmc_in@meta.data %>% select(orig.ident, type)

# Run standard pipeline
pbmc <- CreateSeuratObject(counts = counts, meta.data = metadata, project = "PBMC_Covid")
pbmc <- PercentageFeatureSet(pbmc, pattern = "^MT-", col.name = "percent_mito")
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & percent_mito < 10)

pbmc <- NormalizeData(pbmc, verbose = FALSE)
pbmc <- FindVariableFeatures(pbmc, verbose = FALSE)
pbmc <- ScaleData(pbmc, verbose = FALSE)
pbmc <- RunPCA(pbmc, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:12, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:12, verbose = FALSE)

# Default resolution 0.5 clustering
pbmc <- FindClusters(pbmc, resolution = 0.5, verbose = FALSE)

# Rename idents
new_identities <- c(
  "0" = "T Cells",
  "1" = "CD14+ Monocytes",
  "2" = "T Cells",
  "3" = "B Cells",
  "4" = "CD14+ Monocytes",
  "5" = "NK Cells",
  "6" = "NK Cells",
  "7" = "CD14+ Monocytes",
  "8" = "CD14+ Monocytes",
  "9" = "T Cells",
  "10" = "B Cells",
  "11" = "CD16+ Monocytes",
  "12" = "CD14+ Monocytes"
)
pbmc <- RenameIdents(pbmc, new_identities)
pbmc$cell_type <- Idents(pbmc)
```

---

## Exercise 10.1 — Ribosomal Gene Profiling

### 1. Calculate ribosomal percentage

In human cells, ribosomal proteins start with `RPS` (small subunit) or `RPL` (large subunit).

```{r calc-ribo}
pbmc <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]", col.name = "percent_ribo")
head(pbmc@meta.data)
```

### 2. Scatter plot: Ribo vs. Mito

```{r plot-ribo-mito}
FeatureScatter(pbmc, feature1 = "percent_ribo", feature2 = "percent_mito")
```

### 3. Discussion

- **Correlation**: We observe a slightly negative or neutral correlation between ribosomal and mitochondrial percentages. High mitochondrial content (>10%) is typically a sign of cell death (leakage of cytoplasmic RNA, leaving only organelles like mitochondria behind).
- **Ribosomal Interpretation**: High ribosomal protein gene expression is characteristic of active, healthy, and rapidly growing or translating cells (e.g., active lymphocytes). It does *not* indicate cellular damage or death. Therefore, we filter out cells with high mitochondrial content, but we usually retain cells with high ribosomal content.

---

## Exercise 10.2 — Granular Clustering Resolution

### 1 & 2. Run clustering with resolution = 1.2

```{r high-res-clustering}
# Find clusters at high resolution
pbmc <- FindClusters(pbmc, resolution = 1.2, verbose = FALSE)

# Number of clusters
length(unique(pbmc$seurat_clusters))
table(pbmc$seurat_clusters)
```

By changing the resolution from `0.5` (which generated 13 clusters, labeled 0–12) to `1.2`, we now get 15 clusters.

### 3. Generate high-resolution UMAP

```{r plot-high-res-umap}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```

### 4. Inspect T-cell clusters

Under the default resolution of 0.5, T-cells occupied clusters 0, 2, and 9.
At resolution 1.2, we can see that the T-cell population is split into more granular sub-clusters. This allows researchers to identify CD4+ vs CD8+ T cells, or naive vs memory T cell states, which require higher resolution to resolve.

---

## Exercise 10.3 — NK Cells vs. T Cells Differential Expression

### 1. Run markers comparing NK cells vs. T cells

We reset the identities to the annotated `cell_type` and compare them.

```{r nk-vs-t}
Idents(pbmc) <- "cell_type"

nk_vs_t_markers <- FindMarkers(
  object = pbmc,
  ident.1 = "NK Cells",
  ident.2 = "T Cells",
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)
```

### 2 & 3. Filter and list top 5 NK markers

```{r show-nk-markers}
significant_nk <- nk_vs_t_markers %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < 0.05, avg_log2FC > 0.5) %>%
  arrange(desc(avg_log2FC))

significant_nk %>%
  dplyr::slice_head(n = 5) %>%
  select(gene, avg_log2FC, p_val_adj, pct.1, pct.2)
```

### 4. Discussion

The top marker genes (such as `NKG7`, `GNLY`, `GZMB`, `PRF1`, and `CST7`) encode cytotoxic effector molecules:
- **`GZMB`** (Granzyme B) and **`PRF1`** (Perforin) are responsible for forming pores in target cell membranes and inducing apoptosis.
- **`GNLY`** (Granulysin) has antimicrobial activity and induces apoptosis in infected cells.
- **`NKG7`** is involved in cytotoxic granule exocytosis.
These are classic markers of natural killer (NK) cells and cytotoxic T lymphocytes, confirming that Cluster 6 has cytotoxic effector identity.

---

## Exercise 10.4 — B Cell Antiviral Response in COVID-19

### 1 & 2. Run B Cell COVID-19 vs. Control DGE

```{r b-cell-dge}
# Subset to B Cells only
b_cells <- subset(pbmc, idents = "B Cells")

# Compare Covid vs Control
Idents(b_cells) <- "type"
b_cell_dge <- FindMarkers(
  object = b_cells,
  ident.1 = "Covid",
  ident.2 = "Ctrl",
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)
```

### 3. List top 5 up-regulated genes in B cells

```{r show-b-cell-up}
b_cell_dge %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
  arrange(desc(avg_log2FC)) %>%
  dplyr::slice_head(n = 5) %>%
  select(gene, avg_log2FC, p_val_adj)
```

### 4. Comparison with CD14+ Monocyte response

- In CD14+ Monocytes, the top up-regulated genes in COVID-19 included `ISG15`, `IFITM3`, `LYZ`, and interferon-responsive targets.
- In B Cells, the top up-regulated genes include many of the same interferon-stimulated genes (ISGs), such as **`ISG15`**, **`IFI6`**, and **`IFITM3`**.
- **Conclusion**: Both B cells and CD14+ monocytes exhibit a strong Type I interferon signature (`ISG15`, `IFI6`, `IFITM3`) indicating a systemic antiviral response in blood cells of COVID-19 patients. However, cell-type-specific responses are also present: monocytes express more inflammation-promoting cytokines and chemokines, whereas B cells show activation of pathways involved in antibody production and presentation.
