---
title: "Module 10 - Capstone Starter Notebook"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  data_dir:   "data"
  output_dir: "outputs"
  seed:       42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)

set.seed(params$seed)
```

# Capstone: Clinical and Omics Integration for Survival Prediction

## Your biological question

> Replace this line with one sentence that states your specific biological question.

Example: "Does incorporating gene expression data from post-surgical breast tissue samples improve prediction of 5-year relapse-free survival beyond what is achievable with treatment assignment alone?"

## Author information

```
Name:      [Your name]
Date:      [Analysis date]
Course:    Data Science for Biologists — R Edition
Data used: data/clinical.csv, data/omics.csv
```

---

# Setup

```{r packages}
library(tidyverse)
library(here)
library(survival)
library(survminer)
library(tidymodels)
library(glmnet)
library(limma)
library(scales)

# Create output directory if needed
dir.create(here(params$output_dir), showWarnings = FALSE)
```

---

# Section 1: Data Loading and Validation

This section is required. Every capstone report must demonstrate that the data were loaded and validated correctly.

```{r load-data}
clinical <- read_csv(here(params$data_dir, "clinical.csv"), show_col_types = FALSE)
omics    <- read_csv(here(params$data_dir, "omics.csv"),    show_col_types = FALSE)
```

```{r validate-ids}
# --- Required checks ---
stopifnot(!anyDuplicated(clinical$patientID))
stopifnot(!anyDuplicated(omics$patientID))
stopifnot(setequal(clinical$patientID, omics$patientID))

omics_aligned <- omics %>%
  slice(match(clinical$patientID, patientID))

stopifnot(identical(clinical$patientID, omics_aligned$patientID))
```

```{r dataset-contract}
# Document what you are working with
tibble(
  field   = c("Patients", "Clinical columns", "Expression probes",
               "Missing clinical values", "Missing expression values",
               "Observed events", "Censored observations"),
  value   = c(
    nrow(clinical),
    ncol(clinical),
    ncol(omics) - 1,
    sum(is.na(clinical)),
    sum(is.na(select(omics, -patientID))),
    sum(clinical$status == 1),
    sum(clinical$status == 0)
  )
)
```

---

# Section 2: Exploratory Analysis

This section is required. Provide at least three exploratory plots. Label each with a biological interpretation.

```{r eda-survival-by-treatment}
clinical %>%
  mutate(
    treatment = factor(treatment, levels = c(0, 1),
                       labels = c("Standard of care", "Standard + chemotherapy")),
    status    = factor(status, levels = c(0, 1), labels = c("Censored", "Event"))
  ) %>%
  ggplot(aes(x = treatment, fill = status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Event fraction by treatment group",
    x = NULL, y = "Fraction of patients", fill = "Outcome"
  ) +
  theme_minimal()
```

```{r eda-follow-up}
# Add more EDA here. Ideas:
# - Follow-up time distribution by treatment
# - Expression distribution for a few probes
# - PCA coloured by treatment and outcome
```

---

# Section 3: Kaplan-Meier Analysis

This section is required.

```{r km-analysis}
surv_data <- clinical %>%
  mutate(
    treatment = factor(treatment, levels = c(0, 1),
                       labels = c("standard", "standard_plus_chemo"))
  )

km_fit <- survfit(Surv(time, status) ~ treatment, data = surv_data)

ggsurvplot(
  km_fit,
  data          = surv_data,
  pval          = TRUE,
  conf.int      = TRUE,
  risk.table    = TRUE,
  risk.table.height = 0.25,
  palette       = c("#2a6f97", "#d73027"),
  xlab          = "Time (years)",
  ylab          = "Survival probability",
  title         = "Kaplan-Meier survival curves by treatment group",
  legend.labs   = c("Standard", "Standard + chemo"),
  ggtheme       = theme_minimal()
)
```

**Interpretation:** [Replace with your interpretation. Address the direction of the difference, the log-rank p-value, and at least one limitation of this comparison.]

---

# Section 4: Clinical Baseline Model

This section is required.

```{r clinical-model}
cox_clinical <- coxph(Surv(time, status) ~ treatment, data = surv_data, ties = "efron")
summary(cox_clinical)

# Proportional hazards check
ph_test <- cox.zph(cox_clinical)
print(ph_test)
```

**Interpretation:** [Report the hazard ratio, 95 % CI, and p-value. State whether the proportional hazards assumption was met.]

---

# Section 5: Expression-Aware Analysis

This section is required. At least one analysis must incorporate expression data. Choose one or more of the approaches below and fill in the required interpretation.

## Option A: Differential expression by treatment

```{r limma-treatment}
expression_matrix <- omics_aligned %>% select(-patientID) %>% as.matrix()
rownames(expression_matrix) <- omics_aligned$patientID

sample_metadata <- clinical %>%
  transmute(
    patientID,
    treatment_group = factor(
      if_else(treatment == 1, "standard_plus_chemo", "standard"),
      levels = c("standard", "standard_plus_chemo")
    )
  )

design <- model.matrix(~ treatment_group, data = sample_metadata)
fit    <- lmFit(t(expression_matrix), design)
fit    <- eBayes(fit)

de_results <- topTable(
  fit,
  coef   = "treatment_groupstandard_plus_chemo",
  number = Inf,
  sort.by = "P"
) %>%
  rownames_to_column("probe_id") %>%
  as_tibble() %>%
  mutate(significant = adj.P.Val < 0.05)

tibble(
  total_probes     = nrow(de_results),
  fdr_significant  = sum(de_results$significant)
)
```

**Interpretation:** [Describe the number of significant probes, the direction of effects, and what you cannot conclude without gene annotation.]

## Option B: Penalized Cox with expression predictors

```{r penalized-cox, eval=FALSE}
# Set eval=TRUE when you are ready to run this section.
# Reminder: fit this on the TRAINING set only.

x <- expression_matrix
y <- Surv(clinical$time, clinical$status)

cv_cox <- cv.glmnet(x, y, family = "cox", alpha = 0.5, nfolds = 5)
plot(cv_cox)

# Extract selected probes
selected <- coef(cv_cox, s = cv_cox$lambda.min) %>%
  as.matrix() %>%
  as_tibble(rownames = "probe_id") %>%
  rename(coef = `1`) %>%
  filter(coef != 0) %>%
  arrange(desc(abs(coef)))

selected
```

**Interpretation:** [Describe which probes were selected, their coefficient direction, and the C-index. Explicitly note whether this is in-sample or validated.]

---

# Section 6: Model Comparison

```{r model-comparison}
# Fill in the C-indices from your models above.
# If you did not run penalized Cox, leave the omics row as NA.

tribble(
  ~model,                   ~c_index, ~validation_type,
  "Clinical only (Cox)",    NA_real_, "In-sample",
  "Expression (penalized)", NA_real_, "In-sample — report held-out if available",
  "Combined",               NA_real_, "In-sample — report held-out if available"
)
```

---

# Section 7: Leakage Discussion

This section is required. Answer in writing.

**Question:** In your expression-aware analysis, at which step could patient outcome information have leaked into the feature selection or preprocessing? How did you prevent it?

[Your answer here — 3 to 5 sentences.]

---

# Section 8: Biological Interpretation and Limitations

This section is required.

**Biological interpretation:** [What do your results suggest, biologically? Be specific about effect directions and magnitudes.]

**Limitations:** [List at least four limitations. Consider: sample size, censoring, confounding, feature annotation, in-sample validation, generalisability, clinical translation.]

---

# Section 9: Reproducibility

```{r session-info}
sessionInfo()
```

**Reproducibility note:** To reproduce this analysis, clone the repository, run `renv::restore()`, then run `rmarkdown::render("course/10_capstone/10_capstone_starter.Rmd")` from the project root.
