---
title: "R for Biologists | Module 10 — Capstone: Clinical and Omics Integration for Survival Prediction"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

Can matched clinical and gene expression data be used to understand and predict survival outcomes in breast cancer patients treated with adjuvant chemotherapy?

# Learning Objectives

By the end of the capstone, learners should be able to:

- validate matched clinical and omics data,
- perform exploratory analysis of survival outcomes,
- build clinical-only and expression-aware models,
- evaluate model performance without leakage,
- interpret molecular features cautiously,
- deliver a reproducible report with assumptions and limitations.

# Required Deliverables

Each learner or team must submit:

1. A rendered R Markdown or Quarto report.
2. The source notebook or script files.
3. A short summary of the biological question and result.
4. A limitations section.
5. A reproducibility note describing package versions and how to rerun the analysis.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(survival)
```

# Load and Validate Data

```{r load-data}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE)
omics <- read_csv(here("data", "omics.csv"), show_col_types = FALSE)

stopifnot(!anyDuplicated(clinical$patientID))
stopifnot(!anyDuplicated(omics$patientID))
stopifnot(setequal(clinical$patientID, omics$patientID))
```

Align the expression table to the clinical table explicitly. This protects against accidental row-order assumptions.

```{r align-data}
omics_aligned <- omics %>%
  slice(match(clinical$patientID, patientID))

stopifnot(identical(clinical$patientID, omics_aligned$patientID))

analysis_data <- clinical %>%
  left_join(omics_aligned, by = "patientID")

glimpse(analysis_data[, 1:8])
```

# Dataset Contract

```{r dataset-contract}
contract <- tibble(
  item = c(
    "patients",
    "clinical columns",
    "expression probe columns",
    "missing clinical values",
    "missing expression values",
    "observed events",
    "censored observations"
  ),
  value = c(
    nrow(clinical),
    ncol(clinical),
    ncol(omics) - 1,
    sum(is.na(clinical)),
    sum(is.na(select(omics, -patientID))),
    sum(clinical$status == 1),
    sum(clinical$status == 0)
  )
)

contract
```

# Minimum Analysis Scope

Every capstone report must include:

- a data validation section,
- exploratory plots for treatment, time, and status,
- a Kaplan-Meier analysis by treatment,
- at least one clinical baseline model,
- at least one expression-aware analysis,
- a discussion of leakage risks,
- a discussion of biological interpretation limits.

# Recommended Analysis Path

## 1. Exploratory Analysis

```{r exploratory-plot}
clinical %>%
  mutate(
    treatment = factor(
      treatment,
      levels = c(0, 1),
      labels = c("Standard of care", "Standard + chemotherapy")
    ),
    status = factor(status, levels = c(0, 1), labels = c("Censored", "Event"))
  ) %>%
  ggplot(aes(x = treatment, fill = status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Event fraction by treatment group",
    x = NULL,
    y = "Fraction of patients"
  ) +
  theme_minimal()
```

## 2. Survival Baseline

```{r survival-baseline}
surv_obj <- Surv(time = clinical$time, event = clinical$status)
km_fit <- survfit(surv_obj ~ treatment, data = clinical)

km_fit
```

## 3. Clinical Model

The current clinical table has only treatment, time, and status. The clinical baseline is therefore intentionally simple.

```{r clinical-cox}
cox_treatment <- coxph(Surv(time, status) ~ treatment, data = clinical)
summary(cox_treatment)
```

## 4. Expression-Aware Analysis

A valid expression-aware analysis must avoid selecting genes using the full dataset before evaluation. Feature filtering, feature selection, and normalization decisions must be documented.

Starter idea:

- split patients into training and test sets,
- perform feature filtering inside the training set,
- fit a penalized model or a small pre-specified feature model,
- evaluate on held-out patients,
- report uncertainty and limitations.

# Rubric

| Criterion | Excellent |
|---|---|
| Question | Clear biological question and correct unit of analysis |
| Validation | Identifier, missingness, and outcome checks are explicit |
| Statistics | Survival/censoring handled appropriately |
| Modeling | Data splitting and preprocessing avoid leakage |
| Interpretation | Claims distinguish association, prediction, and mechanism |
| Reproducibility | Report renders from a clean session |
| Communication | Figures and captions are clear and biologically meaningful |

# Common Failure Modes

- Selecting top genes on the full dataset before splitting.
- Treating high expression-prediction performance as causal biology.
- Ignoring censoring and modeling survival time as ordinary regression.
- Reporting only accuracy or p-values.
- Failing to prove that patient IDs align between clinical and omics tables.

# Key Takeaways

- The capstone is the course integration point.
- The main scientific challenge is not running a package; it is making valid analysis decisions.
- Clinical and omics integration requires careful validation, leakage control, and modest interpretation.

# Reproducibility

```{r session-info}
sessionInfo()
```
