---
title: "R for Biologists | Module 7 — Omics Analysis: Differential Expression with limma and Microarray Data"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we turn a high-dimensional expression matrix into defensible biological evidence without mixing up assay types, samples, identifiers, or statistical assumptions?

# Learning Objectives

By the end of this module, learners should be able to:

- distinguish RNA-seq counts, microarray intensities, proteomics intensities, and single-cell matrices,
- validate sample metadata and omics matrix alignment,
- separate raw data, normalized data, transformed data, and model-ready data,
- perform basic omics quality control with missingness, distributions, PCA, and sample outlier checks,
- run a probe-level differential association analysis for continuous expression data,
- apply multiple-testing correction and interpret effect sizes,
- explain why probe identifiers are not automatically gene symbols,
- create a ranked molecular result table for downstream interpretation.

# Prerequisites

- Module 2: data wrangling and validation.
- Module 4: statistical reasoning and multiple testing.
- Module 6: high-dimensional biomedical modeling.
- Packages: `tidyverse`, `here`, `limma`, `scales`.
- Datasets: `data/clinical.csv`, `data/omics.csv`.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(limma)
library(scales)
```

# Omics Data Are Not One Data Type

Different omics assays create different numbers. The method must match the measurement process.

```{r assay-comparison}
tribble(
  ~assay_type, ~typical_values, ~common_question, ~common_modeling_family,
  "Bulk RNA-seq", "Integer read counts", "Which genes change between conditions?", "Negative binomial models, such as DESeq2 or edgeR",
  "Microarray", "Continuous probe intensities", "Which probes differ between conditions?", "Linear models, often limma",
  "Proteomics", "Continuous protein intensities with missingness", "Which proteins differ in abundance?", "Linear models, missingness-aware workflows",
  "Single-cell RNA-seq", "Sparse cell-by-gene counts", "Which cell states or clusters differ?", "Single-cell QC, normalization, clustering, differential testing"
)
```

The bundled `omics.csv` file is a continuous expression matrix with probe IDs as columns. It should not be treated as RNA-seq count data.

# Data Input and Validation

The unit of analysis is the patient. Each patient has one clinical row and one expression row.

```{r load-data}
clinical <- read_csv(here("data", "clinical.csv"), show_col_types = FALSE)
omics <- read_csv(here("data", "omics.csv"), show_col_types = FALSE)

stopifnot(!anyDuplicated(clinical$patientID))
stopifnot(!anyDuplicated(omics$patientID))
stopifnot(base::setequal(clinical$patientID, omics$patientID))

omics_aligned <- omics %>%
  slice(match(clinical$patientID, patientID))

stopifnot(identical(clinical$patientID, omics_aligned$patientID))
```

```{r build-metadata-and-matrix}
sample_metadata <- clinical %>%
  transmute(
    patientID,
    treatment_group = factor(
      if_else(treatment == 1, "standard_plus_chemo", "standard"),
      levels = c("standard", "standard_plus_chemo")
    ),
    follow_up_years = time,
    status_label = factor(
      if_else(status == 1, "event", "censored"),
      levels = c("censored", "event")
    )
  )

expression_matrix <- omics_aligned %>%
  select(-patientID) %>%
  as.matrix()

rownames(expression_matrix) <- omics_aligned$patientID

stopifnot(all(vapply(as.data.frame(expression_matrix), is.numeric, logical(1))))
stopifnot(identical(rownames(expression_matrix), sample_metadata$patientID))

tibble(
  patients = nrow(expression_matrix),
  expression_probes = ncol(expression_matrix),
  missing_expression_values = sum(is.na(expression_matrix)),
  minimum_expression = min(expression_matrix, na.rm = TRUE),
  maximum_expression = max(expression_matrix, na.rm = TRUE)
)
```

# Raw, Normalized, Transformed, and Model-Ready

These terms are often mixed together. Keep them separate.

```{r data-states}
tribble(
  ~state, ~example, ~why_it_matters,
  "Raw", "FASTQ reads, CEL files, or raw count matrices", "Closest to instrument output; often unsuitable for direct modeling",
  "Normalized", "Library-size-adjusted counts or normalized microarray intensities", "Makes samples more comparable under assay-specific assumptions",
  "Transformed", "Log counts, variance-stabilized counts, or scaled intensities", "Improves visualization or model assumptions",
  "Model-ready", "A matrix aligned to metadata after filtering and documented preprocessing", "The exact object used for statistics or machine learning"
)
```

For this course dataset, `omics.csv` is already a processed continuous matrix. We can do QC and linear modeling, but we cannot infer raw RNA-seq library sizes or run count-based differential expression from it.

# Quality Control 1: Missingness and Distributions

Omics QC asks whether samples and probes look comparable enough for the planned analysis.

```{r missingness-qc}
sample_missingness <- tibble(
  patientID = rownames(expression_matrix),
  missing_values = rowSums(is.na(expression_matrix)),
  missing_fraction = missing_values / ncol(expression_matrix)
) %>%
  left_join(sample_metadata, by = "patientID")

sample_missingness %>%
  arrange(desc(missing_fraction)) %>%
  slice_head(n = 10)
```

```{r sample-density}
selected_patients <- sample(rownames(expression_matrix), size = 16)

expression_matrix[selected_patients, , drop = FALSE] %>%
  as_tibble(rownames = "patientID") %>%
  pivot_longer(-patientID, names_to = "probe_id", values_to = "expression") %>%
  ggplot(aes(x = expression, group = patientID)) +
  geom_density(alpha = 0.35, color = "#2a6f97") +
  labs(
    title = "Expression distributions for a subset of patients",
    x = "Processed expression intensity",
    y = "Density"
  ) +
  theme_minimal()
```

If one sample has a radically different distribution, check sample processing, batch, extraction quality, and whether the sample ID was swapped.

# Quality Control 2: PCA

PCA summarizes major sources of expression variation. It does not prove biology by itself; it is a diagnostic view.

```{r pca}
expression_pca <- prcomp(
  expression_matrix,
  center = TRUE,
  scale. = FALSE
)

variance_explained <- expression_pca$sdev^2 / sum(expression_pca$sdev^2)

pca_scores <- as_tibble(expression_pca$x[, 1:5], rownames = "patientID") %>%
  left_join(sample_metadata, by = "patientID")

pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, color = treatment_group, shape = status_label)) +
  geom_point(alpha = 0.75, size = 2) +
  labs(
    title = "PCA of processed expression matrix",
    x = paste0("PC1 (", percent(variance_explained[1], accuracy = 0.1), ")"),
    y = paste0("PC2 (", percent(variance_explained[2], accuracy = 0.1), ")"),
    color = "Treatment",
    shape = "Outcome"
  ) +
  theme_minimal()
```

```{r pca-outliers}
pca_scores %>%
  mutate(distance_from_center = sqrt(PC1^2 + PC2^2 + PC3^2 + PC4^2 + PC5^2)) %>%
  arrange(desc(distance_from_center)) %>%
  select(patientID, treatment_group, status_label, distance_from_center) %>%
  slice_head(n = 10)
```

Large PCA distances are not automatic exclusions. They are a request for investigation.

# Quality Control 3: High-Variance Probe Heatmap

Heatmaps are useful for patterns, but they are easy to overinterpret. This plot uses the most variable probes only as a QC view.

```{r high-variance-heatmap}
probe_variance <- apply(expression_matrix, 2, var, na.rm = TRUE)
top_variable_probes <- names(sort(probe_variance, decreasing = TRUE))[1:30]

heatmap_patients <- sample_metadata %>%
  arrange(treatment_group, status_label, patientID) %>%
  slice(round(seq(1, n(), length.out = 60))) %>%
  pull(patientID)

heatmap_scaled <- scale(expression_matrix[heatmap_patients, top_variable_probes, drop = FALSE])

as_tibble(heatmap_scaled, rownames = "patientID") %>%
  pivot_longer(-patientID, names_to = "probe_id", values_to = "scaled_expression") %>%
  mutate(
    patientID = factor(patientID, levels = heatmap_patients),
    probe_id = factor(probe_id, levels = rev(top_variable_probes))
  ) %>%
  ggplot(aes(x = patientID, y = probe_id, fill = scaled_expression)) +
  geom_tile() +
  scale_fill_gradient2(low = "#4575b4", mid = "white", high = "#d73027", midpoint = 0) +
  labs(
    title = "Top variable probes across selected patients",
    x = "Patients ordered by treatment and outcome",
    y = "Probe ID",
    fill = "Scaled\nexpression"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

# Differential Association for Continuous Expression

For this processed microarray-style matrix, a linear modeling framework is appropriate. The model below estimates the treatment association for every probe.

```{r limma-model}
design <- model.matrix(~ treatment_group, data = sample_metadata)

fit <- lmFit(t(expression_matrix), design)
fit <- eBayes(fit)

limma_results <- topTable(
  fit,
  coef = "treatment_groupstandard_plus_chemo",
  number = Inf,
  sort.by = "P"
) %>%
  rownames_to_column("probe_id") %>%
  as_tibble() %>%
  mutate(
    neg_log10_p = -log10(pmax(P.Value, .Machine$double.xmin)),
    significant = adj.P.Val < 0.05,
    rank_score = sign(logFC) * neg_log10_p
  )

limma_results %>%
  select(probe_id, logFC, AveExpr, t, P.Value, adj.P.Val) %>%
  slice_head(n = 10)
```

`logFC` is the estimated expression difference for `standard_plus_chemo` relative to `standard` in the processed expression scale. It is not a causal chemotherapy effect unless the study design supports that claim.

# Multiple Testing and Effect Size

Omics analyses test many probes at once. A small p-value from one probe is not enough; false discovery control is part of the analysis.

```{r multiple-testing-summary}
tibble(
  total_probes_tested = nrow(limma_results),
  probes_with_raw_p_below_0_05 = sum(limma_results$P.Value < 0.05, na.rm = TRUE),
  probes_with_fdr_below_0_05 = sum(limma_results$adj.P.Val < 0.05, na.rm = TRUE),
  probes_with_fdr_below_0_10 = sum(limma_results$adj.P.Val < 0.10, na.rm = TRUE),
  largest_absolute_effect = max(abs(limma_results$logFC), na.rm = TRUE)
)
```

```{r volcano}
limma_results %>%
  mutate(
    result_group = case_when(
      adj.P.Val < 0.05 & logFC > 0 ~ "higher in standard_plus_chemo",
      adj.P.Val < 0.05 & logFC < 0 ~ "lower in standard_plus_chemo",
      TRUE ~ "not FDR-significant"
    )
  ) %>%
  ggplot(aes(x = logFC, y = neg_log10_p, color = result_group)) +
  geom_point(alpha = 0.65, size = 1.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Probe-level treatment association volcano plot",
    x = "Estimated expression difference",
    y = "-log10(p-value)",
    color = NULL
  ) +
  theme_minimal()
```

# Confounding Check

Before interpreting treatment-associated probes, ask whether treatment is tangled with outcome or follow-up patterns.

```{r confounding-check}
sample_metadata %>%
  count(treatment_group, status_label) %>%
  group_by(treatment_group) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()
```

```{r follow-up-by-treatment}
sample_metadata %>%
  ggplot(aes(x = treatment_group, y = follow_up_years, fill = treatment_group)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE) +
  labs(
    title = "Follow-up time by treatment group",
    x = NULL,
    y = "Follow-up time in years"
  ) +
  theme_minimal()
```

If treatment assignment reflects disease severity, clinical site, assay batch, or follow-up differences, the expression associations may reflect confounding rather than treatment biology.

# Identifier and Annotation Caveat

The expression columns are probe identifiers, such as `200008_s_at`, not automatically gene symbols. Pathway analysis and biological naming require a documented annotation source.

```{r probe-id-examples}
tibble(
  probe_id = colnames(expression_matrix)[1:8],
  learner_warning = "Do not treat this identifier as a gene symbol until it is mapped with a documented annotation source."
)
```

Annotation decisions should record:

- identifier type,
- array or assay platform,
- annotation database and version,
- one-to-many mappings,
- probes that map to no gene or multiple genes.

# Ranked Results for Downstream Interpretation

Many pathway methods start from a ranked list. Here the ranking is real, but it is still a probe-level ranking until annotation is added.

```{r ranked-results}
ranked_probe_results <- limma_results %>%
  arrange(desc(rank_score)) %>%
  select(probe_id, rank_score, logFC, P.Value, adj.P.Val)

ranked_probe_results %>%
  slice_head(n = 12)
```

This replaces the dummy ranked gene list pattern from the legacy omics track. A later pathway module should either map probes to gene IDs with a reproducible annotation source or use a dataset that already provides stable gene identifiers.

# Interpretation Checkpoint

Answer these before reporting an omics result:

1. What is the unit of analysis?
2. What assay type produced the values?
3. Are sample IDs aligned to metadata?
4. Which preprocessing state is being modeled?
5. Which coefficient answers the biological question?
6. How many tests were performed?
7. Which identifier type is being interpreted?
8. What confounding or batch variables could explain the pattern?

# Exercises

1. Repeat the PCA plot using PC2 and PC3. Does the treatment pattern look stronger, weaker, or unchanged?
2. Change the limma design to test `status_label` instead of `treatment_group`. Explain why this is an association with observed event status, not a valid survival analysis.
3. Find the 20 probes with the largest absolute `logFC`. How many are FDR-significant?
4. Write a two-sentence caption for the volcano plot that mentions effect size, multiple testing, and probe identifiers.
5. List the metadata columns you would request from the laboratory before trusting a publication-level differential expression result.

# Common Failure Modes

- Running RNA-seq count methods on processed microarray intensities.
- Joining metadata and expression rows by position instead of validated sample IDs.
- Selecting top probes before train/test splitting in a predictive model.
- Treating FDR-significant probes as causal biology.
- Calling probe IDs gene symbols without annotation.
- Ignoring batch, site, extraction date, RNA quality, or platform version.

# Key Takeaways

- Omics analysis starts with assay identity and sample alignment, not a package call.
- The bundled expression matrix is appropriate for continuous-expression QC and linear modeling.
- Multiple testing correction is mandatory for probe-level scans.
- Ranked molecular results are useful only when their identifier type and annotation limits are explicit.
- Omics findings should be reported as evidence under assumptions, not as automatic mechanisms.

# Memorable Rule

Assay decides model.

# Reproducibility

```{r session-info}
sessionInfo()
```
