---
title: "Module X - Short Descriptive Title"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

State the biological or translational question in one or two sentences.

# Learning Objectives

By the end of this notebook, learners should be able to:

- Explain the core concept.
- Run the analysis in R.
- Interpret the result in biological terms.
- Name the main assumptions and limitations.

# Prerequisites

- Required R packages:
- Required datasets:
- Prior modules:

# Data Input and Validation

```{r data-input}
# Load packages and data here.
# Validate dimensions, identifiers, missingness, and expected column types.
```

# Concept

Explain the concept before introducing package mechanics.

# Worked Example

```{r worked-example}
# Keep each code chunk focused on one analytical step.
```

# Interpretation Checkpoint

Write a short interpretation prompt. Learners should explain what the result means and what it does not mean.

# Common Failure Modes

- Failure mode 1:
- Failure mode 2:
- Failure mode 3:

# Exercises

1. Exercise prompt.
2. Exercise prompt.
3. Exercise prompt.

# Key Takeaways

- Takeaway 1.
- Takeaway 2.
- Takeaway 3.

# Reproducibility

```{r session-info}
sessionInfo()
```
