---
title: "C1: RNA-Seq Differential Expression (DESeq2) - Comprehensive Guide"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

# Introduction to Bulk RNA-Seq Analysis

Welcome to the definitive guide on Differential Expression (DE) analysis. Bulk RNA-Sequencing allows us to take a snapshot of the transcriptome across different biological conditions. Our goal is to identify which genes are **significantly up-regulated or down-regulated** in response to a treatment, disease state, or knockout.

## Why not just use a T-test?
In a standard microarray experiment (or proteomics expression), data is continuous and normally distributed, making t-tests viable. RNA-Seq, however, relies on **read counts** (discrete integers). 
Furthermore, RNA-Seq data suffers from **overdispersion**—the variance between biological replicates is much higher than the mean expression. To accurately model this, we use the **Negative Binomial Distribution**, which is the statistical engine driving `DESeq2`.

## The `airway` Dataset
We will use the `airway` dataset from Bioconductor. This experiment treats airway smooth muscle cells with **dexamethasone** (a synthetic glucocorticoid). We have 4 treated samples and 4 untreated (control) samples.

---

# 1. Setup and Pre-processing

First, we load the required libraries. If you are missing any, `BiocManager::install("PackageName")` is your friend.

```{r setup, message=FALSE, warning=FALSE}
library(DESeq2)
library(airway)
library(tidyverse)
library(pheatmap)      # For heatmaps
library(EnhancedVolcano) # For beautiful volcano plots
library(apeglm)        # For Log2 Fold Change Shrinkage
```

## Exploring the Input Data
`DESeq2` strictly requires **un-normalized raw count data**. Do not feed it TPM, FPKM, or RPKM values. The algorithm handles size-factor normalization internally.

```{r data-prep}
data(airway)
count_matrix <- assay(airway)
metadata <- colData(airway)

# Display a slice of the raw integer counts
head(count_matrix[, 1:4])

# It is critical to define the baseline factor level. 
# We want "Treated vs Untreated", so "Untreated" ("untrt") must be the reference level.
metadata$dex <- factor(metadata$dex)
metadata$dex <- relevel(metadata$dex, ref = "untrt")
```

---

# 2. Creating the DESeq Object and Pre-filtering

We initialize the `DESeqDataSet`. The `design = ~ dex` parameter tells the model that our biological question centers entirely around the `dex` treatment variable.

```{r build-object}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = metadata,
  design = ~ dex 
)

# Pre-filtering: Remove genes with virtually no reads. 
# Keeping them only increases memory usage and penalizes our multiple testing corrections.
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
cat("Genes remaining after pre-filtering:", nrow(dds), "\n")
```

---

# 3. Quality Control (QC): PCA and Sample Distances

Before looking for differential expression, you *must* check if your biological replicates actually group together. If an untreated sample clusters with treated samples, you may have a sample swap!

To do PCA correctly, we must stabilize the variance. Counts are highly skewed; we use a Variance Stabilizing Transformation (VST).

```{r qc-pca}
# Apply Variance Stabilizing Transformation
vsd <- vst(dds, blind = FALSE)

# Plot PCA
plotPCA(vsd, intgroup = "dex") +
  theme_minimal() +
  labs(title = "PCA of Airway RNA-Seq Samples", subtitle = "Treated vs Untreated")
```

*Interpretation:* If PC1 separates your treatment groups and accounts for a high percentage of variance, your experiment worked beautifully!

---

# 4. Running the Differential Expression Pipeline

The generic `DESeq()` function does three critical things under the hood:
1. **Estimate Size Factors:** Normalizes for library depth (so a sample with 50 million reads can be compared to one with 20 million).
2. **Estimate Dispersions:** Calculates the variance at a given mean expression level.
3. **Fit Linear Model:** Uses the Negative Binomial Generalized Linear Model (GLM) and performs the Wald test.

```{r run-deseq, message=FALSE, warning=FALSE}
dds <- DESeq(dds)
res <- results(dds)
summary(res)
```

## The Secret Weapon: Log2 Fold Change (LFC) Shrinkage
Genes with very low counts naturally exhibit massive volatility in Fold Change. A gene going from 1 read to 5 reads is technically a massive fold change, but mathematically noisy. **LFC Shrinkage** shrinks these noisy estimates toward zero, giving you more robust effect sizes.

```{r lfc-shrink}
# Extract the exact name of the coefficient we are testing
resultsNames(dds)  

# Apply apeglm shrinkage
res_shrunken <- lfcShrink(dds, coef = "dex_trt_vs_untrt", type = "apeglm")
```

---

# 5. Visualizing the Biological Impact

Now we convert our results to a data frame and visualize our top hits.

### The MA-Plot
The MA plot shows the Mean Expression (X-axis) against the Log2 Fold Change (Y-axis). Shrinkage is clearly visible here—genes on the far left (low mean) are shrunken to 0.

```{r ma-plot}
DESeq2::plotMA(res_shrunken, ylim = c(-5, 5), main = "MA-Plot (Shrunken)")
```

### The Volcano Plot
Volcano plots evaluate Biological Significance (X-axis) versus Statistical Significance (Y-axis). Let's use `EnhancedVolcano` for publication-ready graphics.

```{r volcano-plot, fig.height=8, fig.width=8}
res_df <- as.data.frame(res_shrunken)

EnhancedVolcano(res_df,
    lab = rownames(res_df), # Using Ensembl IDs here; later we will map to Gene Symbols
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'Dexamethasone Treatment (Treated vs Untreated)',
    pCutoff = 10e-5,
    FCcutoff = 1.5,
    pointSize = 2.0,
    labSize = 3.0)
```

### Expression Heatmap of Top 20 Genes
Let's pull the top 20 most significantly altered genes and see how their normalized counts look across all individual samples.

```{r heatmap}
# Get top 20 genes by adjusted p-value
top_hits <- order(res_df$padj)[1:20]
top_genes <- rownames(res_df)[top_hits]

# Extract their VST normalized expression values
mat <- assay(vsd)[top_genes, ]
# Center and scale (Z-score) the rows for visualization
mat <- mat - rowMeans(mat)

# Create an annotation dataframe to color the top bar by treatment
df_anno <- as.data.frame(colData(vsd)[,"dex", drop=FALSE])

pheatmap(mat, 
         annotation_col = df_anno, 
         main = "Top 20 Differentially Expressed Genes",
         show_colnames = FALSE)
```

# Conclusion & Next Steps
You have successfully imported bulk RNA-Seq counts, corrected for sequencing depth, checked quality with PCA, modeled biological variance with Negative Binomial distributions, and isolated the key regulatory drivers of Dexamethasone treatment. 

**Next in this track:** In C2, we will take this list of significant genes and perform **Gene Set Enrichment Analysis (GSEA)** to figure out exactly which cellular pathways (e.g., Immune Response, Apoptosis) these genes control.
