---
title: "C2 - Pathway Enrichment via GSEA"
output: html_document
---

# C2 - Pathway Enrichment (GSEA)

## Useful References
- [fgsea package](https://bioconductor.org/packages/release/bioc/html/fgsea.html)
- [MSigDB (Molecular Signatures Database)](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp)

## Learning Objectives
In this session we will cover:
- Interpreting a ranked list of genes biologically.
- Running Gene Set Enrichment Analysis.
- Visualizing activated biological pathways.

## Why GSEA?
In lesson C1, we found 500 differentially expressed genes. Biologically, looking at 500 names is useless. We want to know: "Are genes associated with Apoptosis statistically over-represented at the top of my list?" GSEA does exactly that.

## Setup
We use the fast GSEA package `fgsea` and pre-compiled pathways (e.g. KEGG or Hallmark pathways).

```{r setup, message=FALSE, warning=FALSE}
library(fgsea)
library(tidyverse)

# Create a dummy "Ranked List" from DESeq2 log2FoldChange values
set.seed(42)
genes <- paste0("Gene_", 1:1000)
ranked_stats <- setNames(rnorm(1000, 0, 2), genes)
ranked_stats <- sort(ranked_stats, decreasing = TRUE)

# Create a dummy pathway
pathways <- list("Apoptosis" = sample(genes, 50),
                 "Cell_Cycle" = sample(genes, 50))
```

## Running the Test
```{r run-gsea}
# Run fgsea
fgsea_results <- fgsea(pathways = pathways, 
                       stats = ranked_stats, 
                       minSize = 15, 
                       maxSize = 500)

head(fgsea_results)
```

## Visualizing the Enrichment Plot
The classic GSEA Enrichment Plot shows the running sum. If the genes in the pathway are mostly at the top of your list (up-regulated), the line shoots up quickly (Activated Pathway).

```{r plot-gsea}
plotEnrichment(pathways[["Apoptosis"]], ranked_stats) + 
  labs(title = "GSEA Enrichment Plot: Apoptosis")
```
