---
title: "C3 - Single-Cell RNA-Seq Basics"
output: html_document
---

# C3 - Single-Cell RNA-Seq Basics

## Useful References
- [Seurat Vignettes](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
- [Orchestrating Single-Cell Analysis with Bioconductor](https://bioconductor.org/books/release/OSCA/)

## Learning Objectives
In this session we will cover:
- How single-cell data differs from bulk RNA-Seq (sparsity / dropouts).
- Using `Seurat` to create cell objects.
- Doing dimensionality reduction (UMAP) to find distinct cell populations.

## Bulk vs Single-Cell
In bulk RNA-Seq (C1), you blend a million cells into a smoothie and measure the average. Single-cell RNA-Seq measures the exact gene expression of *each individual cell*, allowing you to discover brand new tiny subpopulations of immune cells or neurons that the average "smoothie" hid.

## Setup
We use `Seurat`, the foundational tool for scRNA-Seq in R.
```{r setup, eval=FALSE}
library(Seurat)
library(tidyverse)

# Assuming you loaded a raw sparse matrix from 10X Genomics
pbmc.data <- Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")

# Initialize the Seurat object
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "PBMC_Tutorial", min.cells = 3, min.features = 200)
```

## Quality Control & Normalization
scRNA-Seq cells frequently "die" in the machine. A hallmark of dead cells is high mitochondrial gene percent.

```{r qc, eval=FALSE}
# Find mitochondrial percentage
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Filter out dead cells
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

# Normalize the counts for sequencing depth
pbmc <- NormalizeData(pbmc)
```

## Discovering Cell Populations
Finally, we use UMAP to visualize high-dimensional data in 2D space. Cells that are transcriptionally identical form distinct clusters. 

```{r umap, eval=FALSE}
pbmc <- FindVariableFeatures(pbmc)
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc)

# Find clusters using K-nearest neighbors graph
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# Run UMAP to project it in 2D
pbmc <- RunUMAP(pbmc, dims = 1:10)

DimPlot(pbmc, reduction = "umap")
```
