---
title: "R for Biologists | Module 9 — Reproducible Research: renv, R Markdown, Docker, and Git"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How do we structure and document a data science analysis so that any biologist on the team — or a future version of us — can reproduce every result exactly, from a fresh clone, six months from now?

# Learning Objectives

By the end of this module, learners should be able to:

- create a well-structured RStudio project with conventional directories,
- use `here` to write portable file paths that survive across machines,
- use `renv` to lock R package versions into a reproducible snapshot,
- write a scripted analysis entrypoint that renders all notebooks,
- parameterize an R Markdown report so different datasets or parameters can be run without editing the code,
- apply `.gitignore` strategy for data, results, and rendered outputs,
- explain the purpose of Docker and containerization for computational reproducibility,
- describe how a CI pipeline can render notebooks automatically on every push.

# Prerequisites

- Module 1: R foundations, reading files, project structure.
- Module 2: data wrangling and validation workflows.
- Packages: `here`, `renv`, `rmarkdown`, `tidyverse`.
- A git installation accessible from the command line.

# Setup

```{r packages}
library(tidyverse)
library(here)
library(renv)
```

# The Reproducibility Problem

Imagine a collaborator opens your analysis folder three months after you wrote it. What do they find?

```{r reproducibility-scenarios}
tribble(
  ~scenario,                       ~outcome,
  "Hardcoded absolute path",       "Breaks on any machine that is not yours",
  "Undocumented package versions", "Different behaviour after a CRAN update",
  "Data edited manually in Excel", "No record of what changed, when, or why",
  "Scripts run in unknown order",  "Results cannot be reproduced without guessing",
  "Results committed to git",      "Version control bloated; reproducibility still depends on re-running scripts"
)
```

Reproducibility is not a style preference. It is a quality control requirement. Every finding in a published analysis must be traceable back to documented code, documented data, and documented computational environment.

# Project Directory Structure

A good project layout separates inputs, code, and outputs.

```{r directory-convention}
cat(
  "project/",
  "  data/          <- read-only source data, never edited by hand",
  "    raw/         <- original files, never modified",
  "    processed/   <- outputs of documented cleaning scripts",
  "  R/             <- reusable functions, not scripts that run analysis",
  "    helpers.R",
  "    checks.R",
  "  course/        <- analysis notebooks (this project)",
  "  scripts/       <- entrypoint scripts, e.g. render_all.R",
  "  outputs/       <- figures and tables produced by scripts",
  "  project_detail/<- planning and implementation documentation",
  "  renv.lock      <- pinned package versions",
  "  DESCRIPTION    <- project metadata and declared dependencies",
  "  .gitignore     <- files git should not track",
  "  README.md      <- how to reproduce the analysis from scratch",
  sep = "\n"
)
```

The key rule: **code lives in version control; raw data does not (unless it is small and static); results are always regenerated from code.**

# Portable Paths with `here`

Absolute paths break when the project moves to a different machine, user account, or operating system.

```{r here-demo}
# here() always starts from the project root, wherever R is called from
here("data", "clinical.csv")
here("outputs", "survival_plot.png")
here("course", "08_multiomics_and_survival", "08_multiomics_and_survival.Rmd")
```

```{r here-rules}
tribble(
  ~pattern,                  ~problem,                        ~fix,
  "setwd('/home/user/proj')", "Breaks on every other machine", "Use here() instead",
  "read_csv('../data/x.csv')", "Fails when knitting from a subdirectory", "here('data', 'x.csv')",
  "source('helpers.R')",      "Depends on working directory",  "source(here('R', 'helpers.R'))"
)
```

# Dependency Management with `renv`

R packages change. An analysis that runs today may fail in one year when a dependency updates its API, removes a function, or changes default behaviour.

`renv` solves this by recording the exact version of every installed package into a `renv.lock` file.

```{r renv-workflow}
cat(
  "# 1. Initialise renv for the project (run once)",
  "#    renv::init()",
  "#",
  "# 2. After installing or updating packages, snapshot",
  "#    renv::snapshot()",
  "#",
  "# 3. On a fresh clone or new machine, restore the exact environment",
  "#    renv::restore()",
  "#",
  "# 4. Check which packages are out of sync",
  "#    renv::status()",
  sep = "\n"
)
```

```{r renv-lockfile-structure}
cat(
  "renv.lock records, for each package:",
  "  - Package name",
  "  - Version number",
  "  - Repository (CRAN, Bioconductor, GitHub)",
  "  - SHA hash (for GitHub sources)",
  "",
  "This is enough information for renv::restore() to install",
  "exactly the same versions on a different machine.",
  sep = "\n"
)
```

> **Important:** `renv.lock` should be committed to git. The `renv/library/` directory (the actual installed packages) should be in `.gitignore` because it is large and machine-specific.

# The `.gitignore` Strategy

```{r gitignore-rules}
tribble(
  ~file_type,             ~should_be_tracked, ~reason,
  "R scripts and .Rmd",  "Yes",              "These are the primary reproducibility record",
  "renv.lock",           "Yes",              "Pins package versions",
  "DESCRIPTION",         "Yes",              "Declares project dependencies",
  "README.md",           "Yes",              "Entry point for collaborators",
  "data/raw/",           "Only if small and static", "Large data goes in a data repository",
  "outputs/figures/",    "No",               "Always regenerated from scripts",
  "rendered_notebooks/", "No",               "Always regenerated from scripts",
  ".Rhistory, .RData",   "No",               "Machine-specific session debris",
  "renv/library/",       "No",               "Large; restored by renv::restore()"
)
```

# Scripted Analysis Entrypoint

An entrypoint script lets any collaborator reproduce the full analysis with a single command: `Rscript render_all.R`.

```{r scripted-entrypoint}
cat(
  "# render_all.R — entrypoint for the full course",
  "# Run with: Rscript render_all.R",
  "",
  "library(rmarkdown)",
  "",
  "render_notebook <- function(path, ...) {",
  "  message('Rendering: ', path)",
  "  rmarkdown::render(path, ...)",
  "}",
  "",
  "course_notebooks <- list.files(",
  "  path       = 'course',",
  "  pattern    = '\\\\.Rmd$',",
  "  recursive  = TRUE,",
  "  full.names = TRUE",
  ")",
  "",
  "invisible(lapply(course_notebooks, render_notebook))",
  sep = "\n"
)
```

The render check in `scripts/check_course.R` extends this with pass/fail reporting so failures are immediately visible.

# Parameterized R Markdown Reports

Hard-coding dataset names or analysis parameters inside a notebook means copying and editing the notebook each time you want to run it differently. Parameterized reports avoid this.

```{r parameterized-yaml}
cat(
  "---",
  "title: 'Survival Analysis Report'",
  "output: html_document",
  "params:",
  "  data_path: 'data/clinical.csv'",
  "  treatment_var: 'treatment'",
  "  alpha: 0.05",
  "---",
  "",
  "# In the notebook body, use params$data_path, params$treatment_var, etc.",
  "",
  "# Render with different parameters without editing the file:",
  "# rmarkdown::render(",
  "#   'analysis.Rmd',",
  "#   params = list(data_path = 'data/clinical_holdout.csv', alpha = 0.1)",
  "# )",
  sep = "\n"
)
```

Parameterization is especially useful for rerunning analysis on validation cohorts, sensitivity analyses, or different significance thresholds.

# Data Provenance

Every analysis-ready dataset should have a documented lineage.

```{r provenance-template}
tribble(
  ~field,             ~example,
  "Source file",      "data/raw/study_export_2024-01-15.csv",
  "Downloaded from",  "Internal LIMS system, accessed 2024-01-15",
  "Processing script","scripts/01_clean_clinical.R",
  "Output file",      "data/processed/clinical.csv",
  "Key decisions",    "Excluded 3 patients with missing outcome; capped follow-up at 10 years",
  "Analyst",          "A. Ramirez, 2024-01-16",
  "Last verified",    "2024-03-10, re-ran script after R version update, results identical"
)
```

A data dictionary and provenance note prevent the single most common cause of irreproducibility: not knowing which version of the data was analyzed.

# Version Control Discipline

```{r git-workflow}
cat(
  "Good git habits for scientific analysis:",
  "",
  "  git status          <- always check before committing",
  "  git add -p          <- review every change before staging",
  "  git commit -m '...' <- write a message that explains WHY, not WHAT",
  "  git log --oneline   <- review recent history",
  "",
  "Commit message examples:",
  "  GOOD: 'Fix stratified split: events were leaking into test set'",
  "  GOOD: 'Add Schoenfeld residual plot for PH assumption check'",
  "  BAD:  'Update analysis'",
  "  BAD:  'Fixed stuff'",
  "",
  "Never commit: .RData, large data files, rendered HTML, renv/library/",
  sep = "\n"
)
```

# Docker: Freezing the Entire Computational Environment

`renv` pins R package versions. Docker pins everything else: the operating system, system libraries, R version itself, and all compiled dependencies.

```{r docker-concept}
cat(
  "A Dockerfile for an R analysis typically looks like:",
  "",
  "  FROM rocker/r-ver:4.3.1",
  "",
  "  RUN apt-get update && apt-get install -y \\",
  "      libcurl4-openssl-dev libssl-dev libxml2-dev",
  "",
  "  COPY renv.lock renv.lock",
  "  RUN R -e \"install.packages('renv'); renv::restore()\"",
  "",
  "  COPY . /project",
  "  WORKDIR /project",
  "",
  "  CMD [\"Rscript\", \"render_all.R\"]",
  "",
  "Build: docker build -t my-analysis .",
  "Run:   docker run --rm my-analysis",
  sep = "\n"
)
```

The Rocker Project (`rocker-project.org`) provides versioned R base images for this purpose. Using `rocker/r-ver:4.3.1` guarantees the same R version on every machine.

# Continuous Integration for Rendered Notebooks

A CI pipeline can automatically render all notebooks whenever new code is pushed to the repository. This catches render failures before collaborators see broken notebooks.

```{r ci-concept}
cat(
  "A minimal GitHub Actions workflow (.github/workflows/render.yml):",
  "",
  "  on: [push, pull_request]",
  "",
  "  jobs:",
  "    render:",
  "      runs-on: ubuntu-latest",
  "      steps:",
  "        - uses: actions/checkout@v4",
  "        - uses: r-lib/actions/setup-r@v2",
  "          with:",
  "            r-version: '4.3.1'",
  "        - uses: r-lib/actions/setup-renv@v2",
  "        - name: Render all notebooks",
  "          run: Rscript render_all.R",
  sep = "\n"
)
```

CI render checks are the difference between a course that works and a course that used to work.

# Large Data Management

Omics datasets can be gigabytes. Git is not a data repository.

```{r large-data-strategy}
tribble(
  ~data_size,        ~recommended_approach,
  "< 5 MB",          "Commit directly to the git repository if truly static",
  "5 MB - 1 GB",     "Data repository: Zenodo, Figshare, OSF, or institution data store",
  "> 1 GB",          "Cloud storage with signed URLs; or HPC shared filesystem",
  "All sizes",       "Document download URL, version, and checksum in README or data provenance file"
)
```

For this course, `data/clinical.csv` and `data/omics.csv` are small enough to commit. In a real project with full omics data, only the processed, analysis-ready extract should be tracked.

# Putting It Together: Reproducibility Checklist

This checklist applies to any analysis before sharing or publishing.

```{r reproducibility-checklist}
tribble(
  ~item,                                            ~done,
  "All paths use here() or relative paths",        "[ ]",
  "renv.lock committed and up to date",            "[ ]",
  "Raw data documented with source, date, and version", "[ ]",
  "All notebooks render from a fresh R session",   "[ ]",
  "render_all.R or equivalent entrypoint exists",  "[ ]",
  ".gitignore excludes results, renv/library/, .RData", "[ ]",
  "README explains how to reproduce from a clone", "[ ]",
  "sessionInfo() appears in every notebook",       "[ ]",
  "Random seeds are set and documented",           "[ ]",
  "No manually edited intermediate files",         "[ ]"
) %>%
  rename(Checklist = item, Status = done)
```

# Exercises

1. Run `renv::status()` from the project root. Which packages, if any, are not snapshotted?
2. Create a minimal parameterized `.Rmd` that reads a CSV path from `params$data_path` and prints the number of rows. Render it twice with different `params$data_path` values using `rmarkdown::render()` from the console.
3. Add a new entry to `.gitignore` that excludes any file ending in `_cache/`. Confirm with `git status` that cached render files no longer appear as untracked.
4. Write a `README.md` section titled "Reproducing the analysis" that explains the steps from fresh clone to rendered notebooks. Include the `renv::restore()` and `Rscript render_all.R` commands.
5. Inspect the current `DESCRIPTION` file. What would you add to make it accurately describe the course? Add at least one missing field and explain your choice.

# Common Failure Modes

- Using `setwd()` or absolute paths that work only on the author's machine.
- Installing packages interactively without snapshotting `renv.lock`, so a collaborator gets different versions.
- Committing rendered HTML files instead of only committing the `.Rmd` source.
- Running analysis scripts out of order and forgetting to document the correct order in the README.
- Editing a CSV in Excel between scripts and not recording the change anywhere.
- Writing results into a version-controlled directory so git history fills with binary diffs.

# Key Takeaways

- Reproducibility is a first-class scientific requirement, not a cleanup step.
- `here` eliminates path fragility; `renv` eliminates package fragility; Docker eliminates OS fragility.
- Every analysis needs a documented entrypoint so any collaborator can re-run it in the correct order.
- A CI render check catches failures before collaborators see them.
- The goal is that a new team member can clone the repository, run one command, and get identical results.

# Reproducibility

```{r session-info}
sessionInfo()
```
