---
title: "Module 1 - R Foundations for Biologists"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

set.seed(42)
```

# Biological Question

How can we use core R objects and functions to inspect biological data without losing track of sample identifiers, variable types, and scientific meaning?

# Learning Objectives

By the end of this module, learners should be able to:

- create and inspect basic R objects,
- distinguish numeric, character, logical, and factor variables,
- use vectors and tibbles for biological data,
- write small functions for repeated checks,
- load course data with stable file paths,
- perform simple validation on patient-level data.

# Setup

```{r packages}
library(tidyverse)
library(here)
```

# Objects Represent Biological Things

In R, an object is a named value. In biological analysis, names should reflect the thing being represented.

```{r basic-objects}
gene_symbol <- "ESR1"
expression_value <- 10.8
is_receptor_gene <- TRUE

gene_symbol
expression_value
is_receptor_gene
```

# Vectors

A vector stores multiple values of the same basic type. A gene panel, patient ages, or treatment labels can all be represented as vectors.

```{r vectors}
marker_genes <- c("ESR1", "PGR", "ERBB2", "TP53", "GATA3")
patient_ages <- c(45, 62, 71, 53, 80)
received_chemo <- c(TRUE, FALSE, TRUE, TRUE, FALSE)

length(marker_genes)
mean(patient_ages)
table(received_chemo)
```

# Factors

Categorical biological variables should often be represented as factors, especially when order or reference level matters.

```{r factors}
tumor_stage <- factor(
  c("I", "II", "III", "II", "I"),
  levels = c("I", "II", "III", "IV"),
  ordered = TRUE
)

tumor_stage
summary(tumor_stage)
```

# Tibbles

A tibble is a rectangular table. Most course datasets are tibbles after loading.

```{r tibbles}
toy_clinical <- tibble(
  patient_id = paste0("P", 1:5),
  age = patient_ages,
  tumor_stage = tumor_stage,
  treatment = factor(c("A", "B", "A", "A", "B")),
  event = c(1, 0, 1, 1, 0)
)

toy_clinical
glimpse(toy_clinical)
```

# Functions

Functions make repeated checks explicit. This is important because validation should be rerunnable.

```{r functions}
count_missing <- function(x) {
  sum(is.na(x))
}

count_missing(c(1, 2, NA, 4))
```

Write functions with names that describe the biological or data-quality question.

```{r validation-function}
has_unique_ids <- function(ids) {
  !anyDuplicated(ids)
}

has_unique_ids(toy_clinical$patient_id)
```

# Load a Real Course Dataset

Use `here()` so paths are relative to the project root rather than the current RStudio working directory.

```{r load-metabric}
metabric <- read_csv(
  here("data", "metabric_clinical_and_expression_data.csv"),
  show_col_types = FALSE
)

glimpse(metabric)
```

# Inspect Data Types

Before analysis, check whether each column has the expected type.

```{r inspect-types}
metabric_types <- tibble(
  column = names(metabric),
  type = map_chr(metabric, ~ class(.x)[1]),
  missing_values = map_int(metabric, count_missing)
)

metabric_types
```

# Validate Patient Identifiers

The patient ID column is the anchor for patient-level analysis.

```{r validate-ids}
id_checks <- tibble(
  check = c("no missing patient IDs", "patient IDs are unique"),
  passed = c(
    count_missing(metabric$Patient_ID) == 0,
    has_unique_ids(metabric$Patient_ID)
  )
)

id_checks
```

# Summarize Clinically Meaningful Variables

```{r summarize-clinical}
metabric %>%
  summarise(
    patients = n(),
    median_age = median(Age_at_diagnosis, na.rm = TRUE),
    median_survival_time = median(Survival_time, na.rm = TRUE),
    missing_tumor_size = count_missing(Tumour_size),
    missing_tumor_stage = count_missing(Tumour_stage)
  )
```

# Simple String Work

Biological identifiers are often strings. String operations are useful for gene names, probe IDs, sample IDs, and file names.

```{r string-work}
probe_ids <- names(read_csv(here("data", "omics.csv"), n_max = 0, show_col_types = FALSE))
probe_ids <- probe_ids[probe_ids != "patientID"]

head(probe_ids)
str_detect(probe_ids[1:10], "_at$")
```

# Interpretation Checkpoint

Answer these before moving on:

- Which columns are patient identifiers?
- Which variables are categorical?
- Which variables have missing values?
- Which variables are outcomes?
- Which variables are molecular measurements?

# Common Failure Modes

- Treating character columns as meaningful categories without checking spelling or levels.
- Ignoring missing values because functions return `NA`.
- Using row numbers as sample identifiers.
- Hard-coding absolute file paths that only work on one computer.
- Writing code once interactively but not turning repeated checks into functions.

# Exercises

1. Create a vector containing five gene symbols relevant to your biology area.
2. Convert `ER_status` in METABRIC to a factor and count its levels.
3. Write a function called `fraction_missing()` that returns the fraction of missing values in a vector.
4. Use your function to identify the five METABRIC columns with the highest missingness.
5. Use `here()` to load `data/clinical.csv` and check whether `patientID` is unique.

# Key Takeaways

- R objects should be named after the biological thing they represent.
- Vectors, factors, and tibbles are the basic building blocks for this course.
- Functions make data validation repeatable.
- Stable paths and identifier checks prevent many downstream analysis errors.

# Reproducibility

```{r session-info}
sessionInfo()
```
