---
title: "B1 - Handling Imbalanced Clinical Data"
output: html_document
---

# B1 - Handling Imbalanced Clinical Data

## Useful References
- [Tidymodels: themis package](https://themis.tidymodels.org/)
- [Imbalanced-learn Concepts](https://imbalanced-learn.org/)

## Learning Objectives
In this session we will cover:
- What imbalanced data is and why accuracy is a misleading metric.
- How to implement Downsampling and Upsampling.
- How to generate synthetic data using SMOTE within `tidymodels`.

## The Problem
When predicting a rare disease base or subtype (e.g., 95% Control, 5% Disease), a model that simply guesses "Control" every time will be 95% accurate, but utterly useless.

## Setup
```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(themis) # For imbalance recipes
library(tidyverse)

# Create a highly imbalanced synthetic clinical dataset
set.seed(123)
clinical_data <- tibble(
  Gene_A = rnorm(1000, 5, 2),
  Gene_B = rnorm(1000, 10, 3)
) %>%
  mutate(Disease = factor(ifelse(Gene_A + Gene_B > 22, "Rare", "Normal")))

# Check imbalance
table(clinical_data$Disease)
```

## Tidymodels Recipe with SMOTE
Instead of letting the model see the imbalanced data, we add a step in our preprocessing recipe to synthetically generate minority class points. SMOTE (Synthetic Minority Over-sampling Technique) creates synthetic examples by interpolating between existing minority class points.

```{r smote-recipe}
# 1. Create Data Split
set.seed(42)
split <- initial_split(clinical_data, prop = 0.8, strata = Disease)
train_data <- training(split)

# 2. Build Recipe with SMOTE
imbalance_rec <- recipe(Disease ~ ., data = train_data) %>%
  step_normalize(all_predictors()) %>%
  step_smote(Disease, over_ratio = 1) # Force 1:1 balance

# 3. View the balanced data
balanced_train <- prep(imbalance_rec) %>% juice()
table(balanced_train$Disease)
```

By balancing classes during the recipe phase, any downstream model (like Random Forest or Logistic Regression) will see equal representation and focus on distinguishing the true clinical signals.
