---
title: "B2 - Survival Analysis & Cox Models"
output: html_document
---

# B2 - Survival Analysis & Cox Models

## Useful References
- [Survival Analysis in R (survminer)](https://rpkgs.datanovia.com/survminer/)
- [UCLA Kaplan-Meier Survival Analysis](http://www.sthda.com/english/wiki/survival-analysis-basics)

## Learning Objectives
In this session we will cover:
- What Right-Censoring means in clinical trials.
- How to estimate overall survival using Kaplan-Meier curves.
- How to perform multivariate risk analysis using Cox Proportional Hazards.
- Checking proportional hazards assumptions (Schoenfeld residuals).

## The Problem of Time-to-Event
In standard regression, we predict a continuous number. In classification, we predict a category. Survival Analysis predicts a **time-to-event** (e.g., Overall Survival in months until death). 

Why can't we just use standard linear regression here? Because of **Censoring**.
If a patient leaves a study at month 24, we don't know when they died, only that they survived *at least* 24 months. Survival analysis handles this mathematically without dropping valuable data.

## Setup
```{r setup, message=FALSE, warning=FALSE}
library(survival)   
library(survminer)  
library(tidyverse)  

# We will use the built-in 'lung' cancer dataset
data(lung)

# Clean up status: 1=Death, 0=Censored
lung <- lung %>% 
  mutate(Sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
         Status = ifelse(status == 2, 1, 0)) 
```

## Kaplan-Meier Curves
Kaplan-Meier estimates the survival probability over time.

```{r kaplan-meier}
# Fit the curve
km_fit <- survfit(Surv(time, Status) ~ Sex, data = lung)

# Plot
ggsurvplot(
  km_fit, 
  data = lung, 
  pval = TRUE,
  risk.table = TRUE,
  title = "Kaplan-Meier Survival by Sex",
  xlab = "Time (Days)",
  ylab = "Overall Survival Probability"
)
```

## Cox Proportional Hazards Model
**Cox Regression** calculates a **Hazard Ratio (HR)** for continuous and categorical variables simultaneously.

```{r cox-model}
cox_model <- coxph(Surv(time, Status) ~ age + Sex + ph.ecog, data = lung)
summary(cox_model)

# Forest Plot of Hazard Ratios
ggforest(cox_model, data = lung, main = "Hazard Ratios")
```

## Proportional Hazards Assumption
The model assumes the Hazard Ratio is constant over time. We check this using Schoenfeld Residuals: if p < 0.05, the assumption is violated.

```{r test-ph}
ph_test <- cox.zph(cox_model)
ggcoxzph(ph_test)
```
