---
title: "B3 - High-Dimensional Modeling (Elastic Net)"
output: html_document
---

# B3 - High-Dimensional Modeling

## Useful References
- [Glmnet Vignette](https://glmnet.stanford.edu/articles/glmnet.html)
- [Tidymodels: Penalized Regression](https://www.tidymodels.org/learn/models/pls/)

## Learning Objectives
In this session we will cover:
- Dealing with $p > n$ (more genes than patients) scenarios.
- Regularization (Lasso, Ridge, and Elastic Net).
- Using `tidymodels` to tune hyperparameters.

## The Curse of Dimensionality
When analyzing omics data, you often have 20,000 genes but only 100 patient samples. Standard linear regression breaks here completely. **Regularization** adds mathematical penalties that shrink non-informative genes' coefficients down exactly to zero (Lasso) or severely penalizes them (Ridge).

## Setup
We use the `glmnet` engine inside the `tidymodels` framework.
```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(tidyverse)
# Using a dummy dataset for demonstration where columns >> rows
```

## Defining the Elastic Net Model
Elastic net is a combination of Lasso (shrinks coefficients to 0 for feature selection) and Ridge (handles correlated groups of genes).

```{r elastic-net}
elastic_net_spec <- linear_reg(
  penalty = tune(), # We will search for the perfect penalty amount
  mixture = tune()  # Search for the balance between Lasso and Ridge
) %>% 
  set_engine("glmnet")
```

Rather than plugging in numbers randomly, we build a Tuning Grid and use cross-validation folds to mathematically identify the regularisation parameters that predict clinical outcomes most accurately!
